---
title: "DPMirt: Bayesian Semiparametric Item Response Theory"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  A gateway introduction to the DPMirt package for fitting Bayesian
  semiparametric IRT models using Dirichlet Process Mixture priors, with
  emphasis on the three inferential goals that motivate posterior summary
  selection and the complete model family supported by the package.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DPMirt: Bayesian Semiparametric Item Response Theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(DPMirt)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5,
  fig.align = "center",
  out.width = "85%"
)
set.seed(42)

# Color palette for consistent visualizations across all DPMirt vignettes
palette_dpmirt <- c(
  normal = "#56B4E9",
  dpm    = "#E69F00",
  true   = "gray50",
  pm     = "#E41A1C",
  cb     = "#377EB8",
  gr     = "#4DAF4A"
)

# Helper: find extdata files in installed package OR source tree
find_extdata <- function(filename) {
  path <- system.file("extdata", filename, package = "DPMirt")
  if (nzchar(path) && file.exists(path)) return(path)
  path2 <- file.path("..", "inst", "extdata", filename)
  if (file.exists(path2)) return(path2)
  ""
}

# Guard for data-dependent chunks
has_data <- nzchar(find_extdata("vignette_sim_bimodal.rds")) &&
  nzchar(find_extdata("vignette_fit_rasch_normal_bimodal.rds")) &&
  nzchar(find_extdata("vignette_fit_rasch_dpm_bimodal.rds"))
```


## Introduction

When using Item Response Theory models for educational and psychological
assessment, researchers almost always assume that the latent traits---the
abilities, proficiencies, or attitudes being measured---follow a normal
distribution. This assumption is so deeply embedded in standard IRT software
that it rarely receives explicit scrutiny. Yet the populations we study are
often far from normal.

Consider students preparing for a high-stakes graduation exam: one group has
mastered the curriculum, while another group is still struggling with
foundational concepts. Or consider a clinical sample where most respondents
fall in the typical range, but a meaningful minority exhibits extreme scores.
In these settings, forcing a bell curve onto the latent trait distribution
distorts individual ability estimates, misleads ranking decisions, and
produces inaccurate pictures of the population.

**DPMirt** addresses this problem by replacing the parametric normal prior
with a *Dirichlet Process Mixture* (DPM) prior on the latent trait
distribution. The DPM prior is a Bayesian nonparametric tool that lets the
data inform the shape of the ability distribution---whether it turns out to
be unimodal, bimodal, skewed, or something else entirely---without requiring
the researcher to choose a specific parametric form in advance. The package
wraps the NIMBLE probabilistic programming engine and provides a complete
workflow from data simulation through model fitting, posterior summarization,
and diagnostics.

This vignette introduces the motivation, architecture, and model family of
DPMirt, and provides a road map to the rest of the documentation.


## Motivating Example: Why Normality Matters

### The scenario

Imagine a curriculum-based mathematics assessment administered to a
heterogeneous student population. Advanced-track students have had an
additional year of instruction, while general-track students are still
building fluency. The true ability distribution is genuinely bimodal---two
overlapping groups with distinct centers.

We examine this scenario using pre-computed simulation results shipped with
the package. The data were generated from a Rasch model with 25 items and
200 persons drawn from a bimodal ability distribution (a mixture of two
normal components centered at $\theta = -1.5$ and $\theta = 1.5$). Two models
were fit to these responses: one with a standard normal prior and one with a
DPM prior.

```{r load-motivating-data, eval=has_data}
sim <- readRDS(find_extdata("vignette_sim_bimodal.rds"))
fit_normal <- readRDS(find_extdata("vignette_fit_rasch_normal_bimodal.rds"))
fit_dpm <- readRDS(find_extdata("vignette_fit_rasch_dpm_bimodal.rds"))
```

### The comparison

The figure below overlays three densities: the true generating distribution
(gray), the posterior mean estimates from the normal-prior model (blue), and
the posterior mean estimates from the DPM-prior model (orange).

```{r motivating-figure, eval=has_data, fig.cap="Recovered ability distributions under normal and DPM priors for a bimodal population. The normal prior shrinks estimates toward the center, masking the true bimodality. The DPM prior recovers the two-group structure."}
if (requireNamespace("ggplot2", quietly = TRUE)) {
  library(ggplot2)

  true_theta <- sim$theta
  pm_normal  <- colMeans(fit_normal$theta_samp)
  pm_dpm     <- colMeans(fit_dpm$theta_samp)

  plot_df <- data.frame(
    value  = c(true_theta, pm_normal, pm_dpm),
    source = factor(rep(c("True", "Normal Prior (PM)", "DPM Prior (PM)"),
                        each = length(true_theta)),
                    levels = c("True", "Normal Prior (PM)", "DPM Prior (PM)"))
  )

  fill_vals  <- c("True" = unname(palette_dpmirt["true"]),
                   "Normal Prior (PM)" = unname(palette_dpmirt["normal"]),
                   "DPM Prior (PM)" = unname(palette_dpmirt["dpm"]))

  ggplot(plot_df, aes(x = value, fill = source, color = source)) +
    geom_density(alpha = 0.35, linewidth = 0.8) +
    scale_fill_manual(values = fill_vals) +
    scale_color_manual(values = fill_vals) +
    labs(x = expression(theta), y = "Density",
         title = "Normal vs. DPM Prior: Recovering a Bimodal Population",
         subtitle = "Rasch model, 25 items, 200 persons") +
    theme_minimal() +
    theme(legend.position = "bottom", legend.title = element_blank(),
          plot.title = element_text(face = "bold"))
}
```

### What went wrong with the normal prior?

The normal-prior model has no mechanism to represent bimodality. Its
posterior mean estimates are pulled toward a single central peak---a
phenomenon called *underdispersion* or *overshrinkage*. The two modes of the
true distribution collapse into one, and the tails are artificially thin.

> **Key insight.** When the true ability distribution departs from normality,
> the normal-prior IRT model does not simply produce noisier
> estimates---it produces *systematically biased* estimates. High-ability
> students are pulled downward; low-ability students are pulled upward; and
> the entire empirical distribution function is distorted.

This has practical consequences. If a school district uses a cutoff to
identify students for an enrichment program, the normal-prior estimates will
undercount eligible students from the high-ability group. The DPM prior,
by adapting to the data-generating shape, substantially reduces this bias.


## What DPMirt Does

DPMirt provides a complete toolkit for Bayesian semiparametric IRT,
organized into five functional layers:

```{r architecture-table}
arch_df <- data.frame(
  Layer = c("User Interface", "Step-by-Step",
            "Diagnostics & Visualization", "Simulation", "Integration"),
  Purpose = c("One-step fitting",
              "Fine-grained control over each pipeline stage",
              "MCMC assessment and result visualization",
              "Data generation and loss evaluation",
              "External package interoperability"),
  `Key Functions` = c(
    "dpmirt()",
    "dpmirt_spec(), dpmirt_compile(), dpmirt_sample(), dpmirt_rescale(), dpmirt_estimates()",
    "dpmirt_diagnostics(), plot(), dpmirt_plot_density(), dpmirt_plot_caterpillar()",
    "dpmirt_simulate(), dpmirt_loss()",
    "dpmirt_alpha_prior(), dpmirt_compare()"),
  check.names = FALSE
)
knitr::kable(arch_df, col.names = c("Layer", "Purpose", "Key Functions"),
             caption = "DPMirt package architecture")
```

Most users will interact primarily with the **User Interface** layer. A
single call to `dpmirt()` handles the entire pipeline---from building the
NIMBLE model code, through compilation, MCMC sampling, post-hoc rescaling,
and WAIC computation---returning a fitted model object ready for
summarization and plotting. For users who need more control, the
**Step-by-Step** layer exposes each stage as a separate function, following
a compile-once, sample-many pattern that avoids redundant C++ compilation.


## Three Inferential Goals

One of the central insights motivating DPMirt is that there is no single
"best" point estimate of latent traits. Different applied questions demand
different posterior summaries, and the choice of estimator should be driven
by the inferential goal---not by convention.

Lee & Wind formalized three distinct goals for latent trait estimation:

```{r goals-table}
goals_df <- data.frame(
  Goal = c("1. Individual traits", "2. Ranking", "3. Distribution"),
  Target = c("$\\theta_p$",
             "$R_p = \\text{rank}(\\theta_p)$",
             "$G_N(t) = N^{-1} \\sum I(\\theta_p \\le t)$"),
  Loss = c("MSEL", "MSELR", "ISE / KS"),
  Example = c("What is this student's ability?",
              "Which 5 students need intervention?",
              "What percentage exceed the cutoff?"),
  Estimator = c("Posterior Mean (PM)", "Triple-Goal (GR)",
                "Constrained Bayes (CB) / Triple-Goal (GR)")
)
knitr::kable(goals_df,
  col.names = c("Goal", "Target", "Loss Function",
                "Example Question", "Optimal Estimator"),
  caption = "Three inferential goals for latent trait estimation (Lee & Wind)",
  escape = FALSE)
```

> **Key insight.** No single estimator simultaneously optimizes all three
> goals. The posterior mean (PM) minimizes individual mean squared error but
> produces an underdispersed empirical distribution, distorting rankings and
> distribution estimates. Constrained Bayes (CB) corrects the distribution
> but not the ranks. Only the triple-goal (GR) estimator of Shen and Louis
> (1998) simultaneously targets accurate ranking and distribution recovery.

DPMirt computes all three estimators from stored posterior samples, enabling
researchers to choose the summary that matches their applied question. The
benefit of using a DPM prior compounds with the right estimator choice: the
DPM prior provides a more faithful posterior, from which the CB and GR
estimators can extract more accurate summaries.


## Model Family

DPMirt supports a complete family of unidimensional IRT models, crossing
three item models with two latent trait priors:

```{r model-table}
model_df <- data.frame(
  Model = c("Rasch", "Rasch", "2PL", "2PL", "3PL", "3PL"),
  Prior = c("Normal", "DPM", "Normal", "DPM", "Normal", "DPM"),
  Parameters = c(
    "$\\beta_i$", "$\\beta_i$",
    "$\\beta_i, \\lambda_i$", "$\\beta_i, \\lambda_i$",
    "$\\beta_i, \\lambda_i, \\delta_i$",
    "$\\beta_i, \\lambda_i, \\delta_i$"),
  Parameterizations = c("IRT", "IRT", "IRT, SI", "IRT, SI",
                         "IRT, SI", "IRT, SI"),
  Identification = c(
    "Constrained item*, ability, unconstrained",
    "Constrained item*, unconstrained",
    "Unconstrained*, ability, constrained item",
    "Unconstrained*",
    "Unconstrained*, ability",
    "Unconstrained*")
)
knitr::kable(model_df,
  col.names = c("Model", "Prior", "Item Parameters",
                "Parameterizations", "Identification"),
  caption = "Supported model family (* = default identification strategy)",
  escape = FALSE)
```

- **Parameterizations.** The IRT form uses
  $\text{logit}(\pi_{ip}) = \lambda_i (\theta_p - \beta_i)$; the
  slope-intercept (SI) form uses
  $\text{logit}(\pi_{ip}) = \gamma_i + \lambda_i \theta_p$. Both are
  available for 2PL and 3PL; the Rasch model uses IRT form only.

- **Identification.** IRT models require constraints to resolve location
  (and, for 2PL/3PL, scale) indeterminacy. DPMirt supports constraining
  item parameters, constraining abilities, or leaving the model
  unconstrained with post-hoc rescaling. The unconstrained + rescaling
  approach is the default for 2PL/3PL, following Paganin et al. (2023).

- **DPM prior.** Uses the Chinese Restaurant Process (CRP) representation
  with cluster atoms from a Normal-Inverse-Gamma base measure. The
  concentration parameter $\alpha$ receives a Gamma hyperprior---either the
  default $\text{Gamma}(1, 3)$ or a principled choice via `DPprior`.


## Backbone References

DPMirt builds on three lines of methodological work, each credited here.

**Paganin et al. (2023)** established the computational framework: NIMBLE
model code, custom MCMC samplers, post-hoc rescaling functions, and DP
measure sampling routines. Their unconstrained + rescaling identification
strategy and centered sampler for the SI parameterization form the backbone
of DPMirt's NIMBLE engine. The package wraps this validated codebase in a
user-friendly API with programmatic model code generation.

**Lee & Wind** formalized the three inferential goals---individual trait
estimation, ranking, and distribution recovery---and evaluated PM, CB, and
GR estimators across the full Rasch/2PL/3PL model family under both
parametric and semiparametric priors. A key finding was that test reliability
is the dominant moderator: at high reliability, the combination of DPM prior
with the GR estimator yields substantial gains for ranking and distribution
recovery.

**Lee (2026)** developed the DPprior package for principled specification of
the Gamma hyperprior on $\alpha$. Rather than an arbitrary default,
researchers express beliefs about expected latent subgroups and uncertainty,
and DPprior translates these into calibrated Gamma($a$, $b$) hyperparameters.
This is especially important when sample size is moderate and item quality
is limited, where the prior on $\alpha$ can substantially influence
posterior inference.


## Installation

DPMirt requires NIMBLE as its MCMC backend. NIMBLE in turn requires a
working C++ compiler (Rtools on Windows, Xcode Command Line Tools on macOS,
or `build-essential` on Linux).

```{r installation, eval=FALSE}
# From GitHub (development version)
# install.packages("remotes")
remotes::install_github("joonho112/DPMirt")

# Verify that NIMBLE is working
library(nimble)
nimbleOptions(verbose = FALSE)
```

We recommend also installing these optional packages:

```{r optional-packages, eval=FALSE}
# Visualization (highly recommended)
install.packages("ggplot2")

# Principled alpha elicitation
remotes::install_github("joonho112/DPprior")

# Reliability-targeted simulation
remotes::install_github("joonho112/IRTsimrel")
```

After installation, load the package:

```{r load-dpmirt, eval=FALSE}
library(DPMirt)
```


## Road Map to the Vignettes

DPMirt documentation is organized into two tracks. Applied researchers
follow a progressive path from a quick first fit through the complete model
workflow. Methodological researchers can dive into mathematical and
computational details.

### Applied Researchers Track

```{r applied-track-table}
applied_df <- data.frame(
  Vignette = c("[Quick Start](quick-start.html)",
               "[Models and Workflow](models-and-workflow.html)",
               "[Posterior Summaries](posterior-summaries.html)",
               "[Prior Elicitation](prior-elicitation.html)",
               "[Simulation Study](simulation-study.html)"),
  Purpose = c("Your first IRT fit in 5 minutes",
              "Complete guide to all models and the step-by-step pipeline",
              "PM, CB, GR: when each estimator is optimal",
              "Principled DPM hyperprior selection via DPprior",
              "Replicating the evaluation framework from Lee & Wind"),
  Time = c("5 min", "30--40 min", "25--30 min", "20--25 min", "30--40 min")
)
knitr::kable(applied_df, col.names = c("Vignette", "Purpose", "Time"),
             caption = "Applied researchers track")
```

**Recommended path for new users:**
Quick Start $\rightarrow$ Models and Workflow $\rightarrow$ Posterior
Summaries. The Prior Elicitation and Simulation Study vignettes can be
read in any order after that.

### Methodological Researchers Track

```{r methods-track-table}
methods_df <- data.frame(
  Vignette = c("[Mathematical Foundations](theory-irt-dpm.html)",
               "[NIMBLE Internals](nimble-internals.html)"),
  Purpose = c(
    "IRT model theory, DPM priors, identification, and posterior summary derivations",
    "Custom samplers, compile-once pattern, and advanced NIMBLE configuration"),
  Time = c("45--60 min", "30--40 min")
)
knitr::kable(methods_df, col.names = c("Vignette", "Purpose", "Time"),
             caption = "Methodological researchers track")
```

**Recommended path for methodologists:**
Start with Mathematical Foundations, then consult NIMBLE Internals. The
Applied Track vignettes provide complementary hands-on demonstrations.

### Reading paths by question

- **"I just want to fit a model quickly."**
  $\rightarrow$ [Quick Start](quick-start.html)
- **"Should I use a normal or DPM prior?"**
  $\rightarrow$ [Models and Workflow](models-and-workflow.html), Section 2
- **"Which estimator should I report---PM, CB, or GR?"**
  $\rightarrow$ [Posterior Summaries](posterior-summaries.html)
- **"How do I choose the concentration parameter alpha?"**
  $\rightarrow$ [Prior Elicitation](prior-elicitation.html)
- **"I want to run a simulation study comparing priors."**
  $\rightarrow$ [Simulation Study](simulation-study.html)
- **"How does the DPM-IRT model work mathematically?"**
  $\rightarrow$ [Mathematical Foundations](theory-irt-dpm.html)
- **"I need to customize the NIMBLE sampler configuration."**
  $\rightarrow$ [NIMBLE Internals](nimble-internals.html)


## References

Ghosh, M. (1992). Constrained Bayes estimation with applications.
*Journal of the American Statistical Association*, 87(418), 533--540.

Lee, J. & Wind, S. Targeting toward inferential goals in Bayesian Rasch
models for estimating person-specific latent traits. *OSF Preprint*.
<https://doi.org/10.31219/osf.io/qrw4n>

Lee, J. (2026). Design-conditional prior elicitation for Dirichlet
Process mixtures: A unified framework for cluster counts and weight
control. arXiv preprint arXiv:2602.06301.
https://arxiv.org/abs/2602.06301

Paganin, S., Paciorek, C. J., Wehrhahn, C., Rodríguez, A.,
Rabe-Hesketh, S., & de Valpine, P. (2023). Computational strategies
and estimation performance with Bayesian semiparametric item response
theory models. *Journal of Educational and Behavioral Statistics*,
48(2), 147--188. https://doi.org/10.3102/10769986221136105

Shen, W., & Louis, T. A. (1998). Triple-goal estimates in two-stage
hierarchical models. *Journal of the Royal Statistical Society: Series B*,
60(2), 455--471.


---

*For questions, bug reports, or feature requests, please visit the
[GitHub repository](https://github.com/joonho112/DPMirt).*
