---
title: "The Complete Guide to Models and Workflows"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  A comprehensive guide to every model, prior, parameterization, and
  identification strategy available in DPMirt.  Covers both the one-step
  dpmirt() wrapper and the step-by-step pipeline, with practical advice
  on when to use each workflow.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Complete Guide to Models and Workflows}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(DPMirt)
library(ggplot2)
knitr::opts_chunk$set(
  collapse    = TRUE,
  comment     = "#>",
  fig.width   = 7,
  fig.height  = 5,
  fig.align   = "center",
  out.width   = "85%"
)
set.seed(42)

# DPMirt palette
pal <- list(
  parametric     = "#56B4E9",
  semiparametric = "#E69F00",
  reference      = "gray50",
  accent         = "#009E73",
  warn           = "#D55E00"
)

# Helper: find extdata files in installed package OR source tree
find_extdata <- function(filename) {
  path <- system.file("extdata", filename, package = "DPMirt")
  if (nzchar(path) && file.exists(path)) return(path)
  path2 <- file.path("..", "inst", "extdata", filename)
  if (file.exists(path2)) return(path2)
  ""
}

# Guard: compact pre-computed fixtures must exist for eval'd chunks
has_data <- nzchar(find_extdata("vignette_fit_rasch_normal.rds")) &&
  nzchar(find_extdata("vignette_fit_rasch_dpm.rds"))

has_2pl <- nzchar(find_extdata("vignette_fit_2pl_dpm.rds"))

has_compare <- has_data

has_bimodal <- nzchar(find_extdata("vignette_sim_bimodal.rds")) &&
  nzchar(find_extdata("vignette_fit_rasch_normal_bimodal.rds")) &&
  nzchar(find_extdata("vignette_fit_rasch_dpm_bimodal.rds"))
```

## 1. Overview

DPMirt provides two ways to fit Bayesian IRT models.

**One-step workflow.**
Call `dpmirt()` and get a fully processed `dpmirt_fit` object in one
function call.
Best for standard analyses where the defaults are appropriate.

**Step-by-step workflow.**
Walk through the pipeline explicitly:
`dpmirt_spec()` $\rightarrow$ `dpmirt_compile()` $\rightarrow$
`dpmirt_sample()` $\rightarrow$ `dpmirt_rescale()` $\rightarrow$
`dpmirt_estimates()`.
Best when you need to inspect intermediate objects, run additional
same-session sampling from a compiled model (compile-once, sample-more), or
modify the MCMC configuration before sampling.

For the same model specification and sampling settings, both workflows
produce the same statistical target.


## 2. The One-Step Workflow

### 2.1 Key Arguments

```{r arg-table, echo = FALSE}
arg_tbl <- data.frame(
  Argument    = c("data", "model", "prior", "parameterization",
                   "identification", "niter", "nburnin", "thin",
                   "thin2", "nchains", "seed", "alpha_prior",
                   "mu_K", "confidence", "base_measure", "M", "rescale",
                   "compute_waic", "compute_dp_density", "save_draws",
                   "save_path", "sampler_config"),
  Default     = c("---", '"rasch"', '"normal"', '"irt"',
                   "NULL", "10000", "2000", "1",
                   "NULL", "1", "NULL", "NULL",
                   "NULL", '"medium"', "list(s2_mu=2, ...)", "50", "TRUE",
                   "TRUE", "TRUE", "TRUE", "NULL", "NULL"),
  Description = c(
    "N x I binary response matrix (persons in rows, items in columns)",
    'IRT model: "rasch", "2pl", or "3pl"',
    'Latent trait prior: "normal" or "dpm"',
    '"irt" or "si" (slope-intercept; 2PL/3PL only)',
    "Identification strategy; NULL selects model default",
    "Total MCMC iterations",
    "Burn-in iterations to discard",
    "Thinning interval for item parameters",
    "Thinning for theta (NULL = same as thin)",
    "Number of sequential sampling runs to pool",
    "Random seed for reproducibility",
    "DPM concentration hyperprior: NULL, c(a,b), or DPprior object",
    "Optional expected cluster count for automatic DPprior calibration",
    "Optional DPprior calibration strength when mu_K is set",
    "DPM base measure hyperparameters (Paganin defaults)",
    "CRP truncation level (max clusters)",
    "Apply post-hoc identification rescaling",
    "Compute WAIC for model comparison",
    "Compute DP density summaries for DPM fits",
    "Reserved; keep TRUE because fits retain in-memory posterior draws",
    "Reserved; disk-backed draw storage is not implemented",
    "Optional expert function hook for NIMBLE MCMC configuration"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(arg_tbl, align = c("l", "l", "l"))
```

Several advanced arguments are intentionally narrow in this release.
`sampler_config` accepts an expert function hook for changing the NIMBLE
MCMC configuration. List-based sampler schemas, custom `item_priors`,
`save_draws = FALSE`, and disk-backed `save_path` storage are reserved for a
future release. DP density reconstruction can be expensive for larger DPM
fits; set `compute_dp_density = FALSE` when you only need summaries,
diagnostics, or estimates.

### 2.2 Examples

Each call below takes roughly 1--3 minutes (dominated by NIMBLE
compilation on first use).  We show the code and load compact
pre-computed fixtures with thinned posterior draws.

**Rasch -- Normal prior**

```r
fit_rasch_n <- dpmirt(
  sim$response, model = "rasch", prior = "normal",
  niter = 10000, nburnin = 2000, thin = 32, thin2 = 32, seed = 100
)
```

```{r load-rasch-normal, eval = has_data}
fit_rasch_n <- readRDS(find_extdata("vignette_fit_rasch_normal.rds"))
```

**Rasch -- DPM prior**

```r
fit_rasch_dpm <- dpmirt(
  sim$response, model = "rasch", prior = "dpm",
  niter = 10000, nburnin = 2000, thin = 32, thin2 = 32, seed = 101
)
```

```{r load-rasch-dpm, eval = has_data}
fit_rasch_dpm <- readRDS(find_extdata("vignette_fit_rasch_dpm.rds"))
```

**2PL -- DPM prior**

```r
fit_2pl <- dpmirt(
  sim$response, model = "2pl", prior = "dpm",
  niter = 10000, nburnin = 2000, thin = 32, thin2 = 32, seed = 104
)
```

```{r load-2pl-dpm, eval = has_2pl}
fit_2pl <- readRDS(find_extdata("vignette_fit_2pl_dpm.rds"))
```

The compact RDS fixtures are intended for summaries, diagnostics, and plots.
They retain evenly thinned posterior draws and cannot be resumed because
compiled NIMBLE pointers are valid only in the R session that created them.

The returned `dpmirt_fit` object gives a compact overview when printed,
and `summary()` adds item-level estimates and DPM cluster diagnostics:

```{r print-fit, eval = has_data}
fit_rasch_n
```

```{r summary-fit, eval = has_data}
summary(fit_rasch_dpm)
```


## 3. The Step-by-Step Workflow

### 3.1 Specification

`dpmirt_spec()` translates your model request into a complete NIMBLE
specification: programmatically generated model code, constants, data,
initial values, and monitor lists.

```r
spec <- dpmirt_spec(
  data = sim$response, model = "rasch",
  prior = "dpm", identification = "constrained_item"
)
```

Inspect the returned `dpmirt_spec` object:

```r
spec$code       # nimbleCode object
str(spec$constants)
spec$monitors   # beta, alpha, zi, muTilde, s2Tilde, logprob nodes
spec$monitors2  # eta (thinned separately via thin2)
```

Because the spec is pure R data (no C++ pointers), it can be safely
saved to disk with `saveRDS()` and loaded in a new session.

### 3.2 Compilation

`dpmirt_compile()` feeds the spec into NIMBLE's compiler, producing
C++ objects that run the sampler at native speed.

```r
compiled <- dpmirt_compile(spec)
```

> **Timing warning.**
> Compilation typically takes 30--120 seconds.
> It only needs to be done **once** per model specification.

> **Session-bound caveat.**
> The compiled object contains external C++ pointers and **cannot** be
> serialized with `saveRDS()`.
> If you restart R, re-run `dpmirt_compile()` on the saved spec.

### 3.3 Sampling

`dpmirt_sample()` runs MCMC on the compiled model.
Because compilation is already done, this step is fast and repeatable.

```r
samples <- dpmirt_sample(
  compiled, niter = 10000, nburnin = 2000,
  thin = 1, thin2 = 1, seed = 42
)
```

**Compile-once, sample-more.**
Continue a same-session run without paying the compilation cost again:

```r
samples <- dpmirt_resume(samples, niter_more = 10000)
fit <- dpmirt_rescale(samples)
```

Repeated `dpmirt_sample(compiled, ...)` calls are useful for separate
same-session runs from one compiled object, but by default each call resets
the sampler and the step-by-step API does not automatically pool those
objects. For convenience, `dpmirt()` supports pooled multi-run outputs
directly via the `nchains` argument. Chain provenance is retained in the
returned object metadata, while fully independent chain initialization is
reserved for the dedicated multi-chain sampling upgrade.

### 3.4 Rescaling

`dpmirt_rescale()` applies post-hoc identification rescaling to the
raw posterior samples and returns a `dpmirt_fit` object that can be
passed directly to `dpmirt_estimates()` and other downstream functions. A
freshly created fit can also be passed to `dpmirt_resume()` while its
compiled model reference is still live in the current R session.

```{r rescale-table, echo = FALSE}
rescale_tbl <- data.frame(
  Model     = c("Rasch", "2PL/3PL (IRT)", "2PL/3PL (SI)"),
  Indeterminacy = c("Location only", "Location + scale",
                     "Location + scale"),
  Rescaling = c(
    "Center beta at 0; shift theta accordingly",
    "Center beta, normalize lambda to geom-mean 1, rescale theta",
    "Center gamma via weighted sum, normalize lambda, rescale theta"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(rescale_tbl, align = c("l", "l", "l"))
```

For the Rasch model:

$$
\beta_i^* = \beta_i - \bar{\beta}, \qquad
\theta_j^* = \theta_j - \bar{\beta}
$$

For 2PL/3PL (IRT parameterization):

$$
s = \left(\prod_{i=1}^{I} \lambda_i\right)^{-1/I}, \quad
\beta_i^* = \frac{\beta_i - \bar{\beta}}{s}, \quad
\lambda_i^* = \lambda_i \cdot s, \quad
\theta_j^* = \frac{\theta_j - \bar{\beta}}{s}
$$

```r
fit <- dpmirt_rescale(samples)

# fit is a dpmirt_fit: pass directly to estimates or resume
est <- dpmirt_estimates(fit)
```

> If the model uses `"constrained_item"` or `"constrained_ability"`
> identification, rescaling is a no-op because parameters are already
> identified through in-model constraints. Some requested combinations are
> rejected earlier: DPM models reject `"constrained_ability"`, and 3PL models
> reject `"constrained_item"`.

### 3.5 Estimates

`dpmirt_estimates()` computes person and item point estimates using
three posterior summary methods.

```{r estimator-table, echo = FALSE}
est_tbl <- data.frame(
  Method   = c("PM", "CB", "GR"),
  Name     = c("Posterior Mean",
               "Constrained Bayes (Ghosh, 1992)",
               "Triple-Goal (Shen & Louis, 1998)"),
  Optimizes = c("Individual MSE (Goal 1)",
                "EDF estimation (Goal 3)",
                "Ranking + EDF (Goals 2 & 3)"),
  stringsAsFactors = FALSE
)
knitr::kable(est_tbl, align = c("l", "l", "l"))
```

```{r estimates-demo, eval = has_data}
est <- dpmirt_estimates(fit_rasch_dpm, methods = c("pm", "cb", "gr"))
head(est$theta, 8)
```

```{r item-est-demo, eval = has_data}
head(est$beta, 8)
```


## 4. Model Specifications

All models use a logistic link for
$\pi_{ij} = P(Y_{ij} = 1 \mid \theta_j, \text{item}_i)$.

### 4.1 Rasch Model

$$
\text{logit}(\pi_{ij}) = \theta_j - \beta_i
$$

The Rasch model has a single item parameter --- difficulty $\beta_i$.
The response probability depends only on the difference between person
ability and item difficulty.

**Identification.**
Location indeterminacy only (adding a constant $c$ to all $\theta$ and
$\beta$ leaves the likelihood unchanged).
Default: `"constrained_item"` (mean-centers $\beta$ during MCMC).

**When to use.**
Most parsimonious IRT model.
Assumes equal discrimination and no guessing.
Stable item estimates even with modest sample sizes.

```r
fit_rasch <- dpmirt(data, model = "rasch", prior = "normal")
```

### 4.2 Two-Parameter Logistic (2PL)

Adds a discrimination parameter $\lambda_i$.
Two equivalent parameterizations are available.

**IRT parameterization:**
$$\text{logit}(\pi_{ij}) = \lambda_i(\theta_j - \beta_i)$$

**Slope--intercept (SI) parameterization:**
$$\text{logit}(\pi_{ij}) = \lambda_i \theta_j + \gamma_i$$

where $\gamma_i = -\lambda_i \beta_i$ is an intercept.

```{r param-comparison, echo = FALSE}
param_tbl <- data.frame(
  Feature        = c("Formula", "Item location", "Interpretation",
                      "Posterior correlation", "Centered sampler",
                      "Default identification"),
  IRT            = c("$\\lambda_i(\\theta_j - \\beta_i)$",
                      "$\\beta_i$ (difficulty)",
                      "$\\beta_i$ on theta scale",
                      "Higher ($\\lambda$--$\\beta$)",
                      "Not applicable",
                      "unconstrained"),
  SI             = c("$\\lambda_i \\theta_j + \\gamma_i$",
                      "$\\gamma_i$ (intercept)",
                      "Baseline log-odds at $\\theta = 0$",
                      "Lower ($\\lambda$--$\\gamma$)",
                      "Auto-enabled",
                      "unconstrained"),
  stringsAsFactors = FALSE
)
knitr::kable(param_tbl, align = c("l", "l", "l"))
```

The SI parameterization reduces posterior correlations between
discrimination and intercept, improving MCMC mixing.
DPMirt automatically enables a centered sampler for SI (Paganin et al.,
2023).

**Identification.**
Both location and scale indeterminacy.
Default: `"unconstrained"` + post-hoc rescaling.

```r
fit_2pl_irt <- dpmirt(data, model = "2pl", prior = "dpm")
fit_2pl_si  <- dpmirt(data, model = "2pl", prior = "dpm",
                       parameterization = "si")
```

### 4.3 Three-Parameter Logistic (3PL)

Adds a lower-asymptote (guessing) parameter $\delta_i \in (0, 1)$:

$$
\pi_{ij} = \delta_i + (1 - \delta_i)\,\text{logistic}\!\big(\lambda_i(\theta_j - \beta_i)\big)
$$

DPMirt places a $\text{Beta}(4, 12)$ prior on each $\delta_i$
(prior mean $\approx 0.25$, concentrating mass in 0.1--0.4).
The item-level `delta` parameters are included in summaries, coefficients,
draw extraction, and applicable plots.

**Practical recommendations:**

- Use longer chains (20,000+ iterations, 5,000+ burn-in).
- Most appropriate for multiple-choice items with genuine guessing.
- Both IRT and SI parameterizations are supported.
- Use the default `"unconstrained"` identification or
  `"constrained_ability"` for Normal-prior models; `"constrained_item"` is
  not implemented for 3PL.
- For simulation, 3PL uses DPMirt's fallback generator and does not target
  `target_rho`; reliability targeting via IRTsimrel is currently Rasch/2PL
  only.

```r
fit_3pl <- dpmirt(
  data, model = "3pl", prior = "dpm",
  niter = 20000, nburnin = 5000, seed = 400
)
```


## 5. Priors: Parametric vs Semiparametric

### 5.1 Normal (Parametric) Prior

$$
\theta_j \sim N(\mu, \sigma^2), \qquad
\mu \sim N(0, 3), \qquad
\sigma^2 \sim \text{Inv-Gamma}(2.01, 1.01)
$$

Standard assumption in most IRT software.  Works well when the ability
distribution is approximately unimodal and symmetric.

### 5.2 DPM (Semiparametric) Prior

The DPM prior uses the Chinese Restaurant Process (CRP) representation:

$$
z_j \mid \alpha \sim \text{CRP}(\alpha), \qquad
\theta_j \mid z_j \sim N\!\big(\tilde{\mu}_{z_j},\; \tilde{\sigma}^2_{z_j}\big)
$$

Each cluster $m$ draws from the base measure $G_0$:

$$
\tilde{\mu}_m \sim N(0, \sigma^2_\mu), \qquad
\tilde{\sigma}^2_m \sim \text{Inv-Gamma}(\nu_1, \nu_2)
$$

The concentration parameter $\alpha$ controls the expected number of
clusters.
By default DPMirt uses the conservative $\text{Gamma}(1, 3)$ hyperprior.
You can also pass `alpha_prior = c(a, b)` or a `DPprior_fit` /
`dpmirt_alpha_prior()` result. In the one-step wrapper, setting `mu_K` and
optionally `confidence` asks DPMirt to call DPprior automatically when
installed; if DPprior is unavailable, it falls back to $\text{Gamma}(1, 3)$.

### 5.3 Hyperparameter Defaults

```{r hyper-table, echo = FALSE}
hyper_tbl <- data.frame(
  Parameter   = c("$\\sigma^2_\\mu$", "$\\nu_1$", "$\\nu_2$",
                   "$M$", "$\\alpha$"),
  Default     = c("2", "2.01", "1.01", "50", "Gamma(1, 3)"),
  Description = c(
    "Base measure: cluster mean prior variance",
    "Base measure: Inv-Gamma shape for cluster variance",
    "Base measure: Inv-Gamma rate for cluster variance",
    "CRP truncation level (maximum clusters)",
    "Concentration parameter prior (Paganin default)"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(hyper_tbl, align = c("l", "l", "l"))
```

These can be modified via `base_measure`:

```r
fit <- dpmirt(data, model = "rasch", prior = "dpm",
              base_measure = list(s2_mu = 4, nu1 = 2.5, nu2 = 1.5),
              M = 100)
```

### 5.4 When to Choose the DPM Prior

- **Non-normal populations.** Heterogeneous test-taker groups (e.g.,
  mixing advanced and introductory students).
- **Multimodal distributions.** Floor/ceiling effects or distinct
  subgroups.
- **Exploratory analysis.** When you have little prior information
  about the ability distribution shape.
- **Shrinkage calibration.** The DPM prior produces less severe
  shrinkage when the true distribution is non-Normal.

When the true distribution is approximately Normal, the DPM often
concentrates posterior mass on few effective clusters and can reproduce
Normal-prior behavior, though it still costs more computation than the
parametric model.

### 5.5 Normal vs. DPM: When Normality Fails

The guidelines above are easier to appreciate with a concrete example.
The compact pre-computed bimodal simulation ships with the package: 200
persons drawn from a two-group mixture ($\theta \sim 0.5\,N(-1.5, 0.5) +
0.5\,N(1.5, 0.5)$) assessed on 25 Rasch items ($\bar{w} \approx 0.8$).

```{r load-bimodal, eval = has_bimodal}
sim_bm    <- readRDS(find_extdata("vignette_sim_bimodal.rds"))
fit_bm_n  <- readRDS(find_extdata("vignette_fit_rasch_normal_bimodal.rds"))
fit_bm_d  <- readRDS(find_extdata("vignette_fit_rasch_dpm_bimodal.rds"))
```

WAIC provides a model-level comparison:

```{r compare-bimodal, eval = has_bimodal}
dpmirt_compare(fit_bm_n, fit_bm_d)
```

In this example the two WAIC values are nearly identical ($|\Delta\text{WAIC}| < 2$),
which is expected: WAIC measures predictive accuracy for the *binary
responses*, and both priors predict item responses equally well.
The distributional question is therefore separate from the predictive WAIC
question. Posterior-mean density plots are a useful first visual check, but
posterior means can still be shrunk under either prior; distribution recovery
should be checked with CB/GR summaries and task-specific losses when truth is
available.

```{r bimodal-density, eval = has_bimodal, fig.height = 5, fig.cap = "True bimodal density (gray) versus posterior mean densities under Normal (blue) and DPM (orange) priors. Posterior means are useful diagnostics but can understate distributional differences."}
true_theta <- sim_bm$theta
pm_normal  <- colMeans(fit_bm_n$theta_samp)
pm_dpm     <- colMeans(fit_bm_d$theta_samp)

df_bm <- data.frame(
  value  = c(true_theta, pm_normal, pm_dpm),
  source = factor(rep(c("True", "Normal (PM)", "DPM (PM)"),
                      each = length(true_theta)),
                  levels = c("True", "Normal (PM)", "DPM (PM)"))
)

ggplot(df_bm, aes(x = value, fill = source, colour = source)) +
  geom_density(alpha = 0.25, linewidth = 0.8) +
  scale_fill_manual(values = c("True"         = pal$reference,
                                "Normal (PM)" = pal$parametric,
                                "DPM (PM)"    = pal$semiparametric)) +
  scale_colour_manual(values = c("True"         = pal$reference,
                                  "Normal (PM)" = pal$parametric,
                                  "DPM (PM)"    = pal$semiparametric)) +
  labs(title = "Recovering a Bimodal Population",
       x = expression(theta), y = "Density") +
  theme_bw() +
  theme(legend.position = "top", legend.title = element_blank())
```

> **Take-away.** When the true distribution departs from normality, the DPM
> prior supplies flexibility that the Normal prior cannot. WAIC may not
> detect this distinction because it evaluates item-response prediction, not
> ability-distribution recovery; posterior-mean densities are only a first
> diagnostic.
> For practical guidance on choosing the posterior summary that best
> serves your inferential goal, see
> [Posterior Summaries](posterior-summaries.html).

**Practical recommendation.** For applied researchers who suspect
non-normality, fit both priors and compare them using (1) WAIC for
item-response prediction and (2) PM/CB/GR density plots for distributional
recovery. In simulations where truth is known, use `dpmirt_loss()` to check
whether the apparent distributional gain also improves the loss tied to your
inferential goal.


## 6. Identification Strategies

```{r id-table, echo = FALSE}
id_tbl <- data.frame(
  Strategy  = c("constrained_ability", "constrained_item",
                 "unconstrained"),
  Rasch     = c("Yes (Normal only)", "Yes (default)", "Yes"),
  `2PL`     = c("Yes (Normal only)", "Yes", "Yes (default)"),
  `3PL`     = c("Yes (Normal only)", "No", "Yes (default)"),
  How       = c(
    "Fix $\\theta \\sim N(0, 1)$; no hyperparameters",
    "Mean-center $\\beta$ during MCMC (+ geom-mean $\\lambda = 1$ for 2PL)",
    "No constraints; post-hoc rescaling via dpmirt_rescale()"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(id_tbl,
  col.names = c("Strategy", "Rasch", "2PL", "3PL", "How It Works"),
  align = c("l", "c", "c", "c", "l"))
```

> **Compatibility rules.**
> The unconstrained + post-hoc rescaling approach yields the most efficient
> MCMC sampler for DPM-IRT models. `constrained_ability` is incompatible
> with the DPM prior, and `constrained_item` is not implemented for 3PL.
> DPMirt raises an error for these combinations instead of silently changing
> the requested model.

```r
# Rasch with unconstrained + post-hoc rescaling
fit <- dpmirt(data, model = "rasch", prior = "normal",
              identification = "unconstrained")

# 2PL with in-model constrained_item centering
fit <- dpmirt(data, model = "2pl", prior = "normal",
              identification = "constrained_item")
```

**How to choose.**
For DPM models, use the defaults (`"constrained_item"` for Rasch,
`"unconstrained"` for 2PL/3PL).
For Normal-prior Rasch/2PL models, all three strategies are available.
For Normal-prior 3PL, use `"unconstrained"` or `"constrained_ability"`.
The `"constrained_ability"` strategy gives the strongest $N(0,1)$
shrinkage.


## 7. Chain Continuation

When an initial run needs more iterations, `dpmirt_resume()` continues
the sampler from its current state without recompilation.

```r
# Initial fit
fit <- dpmirt(data, model = "rasch", prior = "dpm",
              niter = 5000, nburnin = 1000, seed = 42)

# Trace looks non-stationary...
plot(fit, type = "trace")

# Resume: add 10,000 more iterations (no burn-in needed)
resumed_samples <- dpmirt_resume(fit, niter_more = 10000)
resumed_fit <- dpmirt_rescale(resumed_samples)
```

With the step-by-step workflow:

```r
samples <- dpmirt_sample(compiled, niter = 5000, nburnin = 1000)
resumed_samples <- dpmirt_resume(samples, niter_more = 10000)
resumed_fit <- dpmirt_rescale(resumed_samples)
```

> `dpmirt_resume()` requires the compiled model's C++ pointers to be
> alive in the same R session. Saved RDS fits and fixtures cannot restore
> those pointers; if they are expired or absent, recompile from the saved
> specification and start a new sampling run.

The returned `dpmirt_samples` object can then be passed through
`dpmirt_rescale()` (which returns a `dpmirt_fit`) and `dpmirt_estimates()`.


## 8. Model Comparison with WAIC

When available, each `dpmirt_fit` stores a WAIC
(Watanabe--Akaike Information Criterion) value computed during sampling.
The current implementation uses NIMBLE's conditional online WAIC with its
default data-node grouping; explicit person-level WAIC grouping is reserved
for a later model-comparison upgrade.
For chain-labeled multi-run fits, the top-level `fit$waic` is currently
the mean of per-run WAIC values; `dpmirt_diagnostics(fit)$waic_by_chain`
and `dpmirt_diagnostics(fit)$waic_aggregation` preserve that provenance.
For single-run fits with finite comparable WAIC, lower values indicate better
expected out-of-sample item-response prediction. For `mean_of_chain_waic`
provenance, inspect `waic_by_chain` and `waic_aggregation` and interpret
rankings cautiously rather than as pooled posterior WAIC.

```r
comparison <- dpmirt_compare(fit_rasch_n, fit_rasch_dpm)
comparison
```

```{r compare-demo, eval = has_compare}
comparison <- dpmirt_compare(fit_rasch_n, fit_rasch_dpm)
knitr::kable(comparison, digits = 2, row.names = FALSE)
```

The `delta_waic` column shows the difference from the best model.
You can compare any number of fits:

```r
dpmirt_compare(fit_rasch_n, fit_rasch_dpm, fit_2pl)
```

> **Interpreting WAIC differences.**
>
> - $|\Delta\text{WAIC}| < 2$: Models essentially equivalent.
> - $2 < |\Delta\text{WAIC}| < 10$: Moderate preference.
> - $|\Delta\text{WAIC}| > 10$: Strong preference.


## 9. Data Formats

### Matrix Format (Default)

Binary matrix with persons in rows and items in columns.
Missing responses coded as `NA`.

```{r matrix-format, eval = FALSE}
head(sim$response[, 1:6])
#>      [,1] [,2] [,3] [,4] [,5] [,6]
#> [1,]    1    0    1    1    0    1
#> [2,]    0    1    0    1    1    0
```

### Long Format

A data.frame with columns for person, item, and response.
DPMirt detects the format automatically and converts internally.

```{r long-format, eval = FALSE}
head(response_long)
#>   person item response
#> 1      1    1        1
#> 2      1    2        0
```

```r
fit <- dpmirt(response_long, model = "rasch", prior = "normal")
```

> If the long-to-matrix conversion produces a very large or sparse
> matrix, verify dimensions with `dpmirt_spec()` before running the
> full pipeline.


## 10. Visualizations

DPMirt provides 12 plot types via `plot(fit, type = ...)`.
When **ggplot2** is installed it is used automatically; otherwise base R
graphics are produced.

```{r plot-types-table, echo = FALSE}
plot_tbl <- data.frame(
  Type        = c("density", "items", "trace", "clusters",
                   "dp_density", "icc", "wright_map",
                   "parameter_trace", "caterpillar",
                   "density_compare", "info", "pp_check"),
  Description = c(
    "Kernel density of posterior mean theta",
    "Item parameter estimates with posterior intervals",
    "Log-likelihood trace (first-pass visual mixing check)",
    "Active cluster count trace (DPM only)",
    "DP mixture density with credible band (DPM only; can be expensive)",
    "Item Characteristic Curves",
    "Wright map: items and persons on common logit scale",
    "Trace plots for individual parameters",
    "Caterpillar plot of sorted estimates with CIs",
    "Overlaid densities from two fits",
    "Test and item information functions (2PL/3PL)",
    "Posterior predictive check"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(plot_tbl, align = c("l", "l"))
```

```{r plot-density-guide, eval = has_data, fig.height = 4, fig.cap = "Posterior density of person abilities."}
plot(fit_rasch_dpm, type = "density")
```

```{r plot-items-guide, eval = has_data, fig.height = 4, fig.cap = "Item difficulty estimates with posterior intervals."}
plot(fit_rasch_dpm, type = "items")
```

```{r plot-trace-guide, eval = has_data, fig.height = 4, fig.cap = "Log-likelihood trace."}
plot(fit_rasch_dpm, type = "trace")
```

```{r plot-icc-guide, eval = has_data, fig.height = 5, fig.cap = "Item Characteristic Curves for items 1--6."}
plot(fit_rasch_dpm, type = "icc", items = 1:6)
```

```{r plot-wright-guide, eval = has_data, fig.height = 5, fig.cap = "Wright map."}
plot(fit_rasch_dpm, type = "wright_map")
```


## 11. Summary and What's Next?

This vignette covered the full model and workflow space in DPMirt:

- **Two workflow modes**: one-step `dpmirt()` and the step-by-step
  pipeline for compile-once sampling while compiled objects are live.
- **Three IRT models**: Rasch, 2PL, and 3PL, each with Normal or DPM
  priors.
- **Two parameterizations** for 2PL/3PL: IRT and slope--intercept.
- **Three identification strategies** with explicit compatibility limits:
  DPM rejects constrained_ability and 3PL rejects constrained_item.
- **Chain continuation** with `dpmirt_resume()` when the compiled model
  pointer is still live.
- **Model comparison** with WAIC via `dpmirt_compare()`.
- **Twelve built-in plot types** for posterior visualization.

```{r next-table, echo = FALSE}
next_tbl <- data.frame(
  Topic    = c("Choosing an estimator",
               "Setting DPM priors",
               "NIMBLE internals"),
  Vignette = c("[Posterior Summaries](posterior-summaries.html)",
               "[Prior Elicitation](prior-elicitation.html)",
               "[NIMBLE Internals](nimble-internals.html)"),
  What     = c(
    "In-depth PM vs CB vs GR comparison; shrinkage diagnostics; loss evaluation",
    "Optional DPprior calibration of alpha; Gamma(1, 3) fallback",
    "Custom sampler hooks, NIMBLE compilation model, advanced tuning"
  ),
  stringsAsFactors = FALSE
)
knitr::kable(next_tbl, align = c("l", "l", "l"))
```
