---
title: "Posterior Summary Methods: Matching Estimators to Inferential Goals"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  How to choose among posterior mean (PM), constrained Bayes (CB), and
  triple-goal (GR) estimators depending on whether the goal is individual
  scoring, ranking, or distribution recovery.  Tied closely to the APM
  manuscript (Lee & Wind).
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Posterior Summary Methods: Matching Estimators to Inferential Goals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(DPMirt)
library(ggplot2)
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width  = 7,
  fig.height = 5,
  fig.align  = "center",
  out.width  = "85%"
)
set.seed(42)

# DPMirt palette for all figures
palette_dpmirt <- c(
  normal = "#56B4E9",
  dpm    = "#E69F00",
  true   = "gray50",
  pm     = "#E41A1C",
  cb     = "#377EB8",
  gr     = "#4DAF4A"
)

# Helper: find extdata files in installed package OR source tree
find_extdata <- function(filename) {
  path <- system.file("extdata", filename, package = "DPMirt")
  if (nzchar(path) && file.exists(path)) return(path)
  path2 <- file.path("..", "inst", "extdata", filename)
  if (file.exists(path2)) return(path2)
  ""
}

# Guard: pre-computed data available?
has_data <- nzchar(find_extdata("vignette_estimates_comparison.rds"))
has_lowrel_data <- nzchar(find_extdata("vignette_estimates_comparison_lowrel.rds"))
has_loss_data <- nzchar(find_extdata("vignette_loss_results.rds"))
```

## 1. Overview

A Bayesian IRT model delivers a full posterior distribution for every person
parameter $\theta_p$.
Collapsing each posterior to a single point estimate is unavoidable for
operational decisions -- but *which* point estimate should you report?

The answer depends on **what you plan to do with the numbers**.
This vignette introduces three posterior summary methods shipped with
DPMirt --- posterior mean (PM), constrained Bayes (CB), and triple-goal
(GR) --- and shows how to match each one to the inferential goal at hand.
The discussion tracks the framework developed in Lee & Wind (APM
manuscript) and the foundational theory of Louis (1984), Ghosh (1992),
and Shen & Louis (1998).

> **Key take-away.**
> No single set of point estimates simultaneously minimises all three loss
> functions.
> Choosing an estimator is choosing a loss function; choosing a loss
> function is choosing an inferential goal.


## 2. Three Inferential Goals

Bayesian decision theory tells us that the optimal point estimate minimises
the expected posterior loss.
Different loss functions lead to different estimators.
The table below maps three common inferential goals in criterion-based
assessment (CBA) to the loss function each implies and the estimator that
minimises it.

```{r goals-table, echo=FALSE}
goals_df <- data.frame(
  Goal = c(
    "1. Individual traits",
    "2. Ranking",
    "3. Distribution recovery"
  ),
  Target = c(
    "True $\\theta_p$",
    "True rank $R_p$",
    "True EDF $G_N(t)$"
  ),
  Loss = c(
    "MSEL: $\\frac{1}{N}\\sum(\\hat{\\theta}_p - \\theta_p)^2$",
    "MSELR: $\\frac{1}{N}\\sum(\\hat{R}_p/N - R_p/N)^2$",
    "KS: $\\sup_t \\lvert\\hat{G}_N(t) - G_N(t)\\rvert$"
  ),
  Example = c(
    "What is this student's ability?",
    "Which 5 students need intervention?",
    "What proportion scored below the cutoff?"
  ),
  Estimator = c(
    "Posterior Mean (PM)",
    "Triple-Goal (GR)",
    "Triple-Goal (GR)"
  )
)
knitr::kable(
  goals_df,
  col.names = c("Goal", "Target", "Loss Function",
                "Example Question", "Optimal Estimator"),
  caption = "Three inferential goals and their optimal posterior summaries."
)
```

### A concrete CBA scenario

Consider a universal screening programme in which 200 elementary students
complete a 25-item mathematics assessment.
A school psychologist might ask three distinct questions about the same
data:

1. **Individual scoring.** "What is Student 47's latent ability?" --
   Report the PM estimate and its credible interval.
2. **Ranking.** "Which five students fall in the bottom 5% and need Tier 2
   support?" -- Report the GR estimates and select the five lowest.
3. **Distribution recovery.** "What proportion of the cohort falls below
   the state proficiency cutoff $t = -0.5$?" -- Compute the empirical
   distribution of GR estimates and evaluate $\hat{G}_N(-0.5)$.

Each question calls for a different estimator.  Using PM for all three
preserves the ranking, but the distribution-level summary is distorted
by shrinkage.

> **"No single set of point estimates simultaneously minimises all three**
> **loss functions."**
> --- Lee & Wind (APM manuscript)


## 3. The Shrinkage Problem

### Why posterior means compress the distribution

The posterior mean for person $p$ under an exchangeable hierarchical model
can be written as a weighted combination of the likelihood-based estimate
and the prior mean:

$$
\hat{\theta}_p^{\text{PM}} \;=\; w_p\,\hat{\theta}_p^{\text{ML}}
  \;+\; (1 - w_p)\,\mu_\theta
$$

where the shrinkage weight is

$$
w_p \;=\; \frac{\sigma^2_\theta}{\sigma^2_\theta + \text{se}(\hat{\theta}_p)^2}.
$$

For every person, $0 < w_p < 1$, so each PM estimate is pulled toward the
grand mean.
The direct consequence is that the variance of the PM estimates is
smaller than the variance of the true abilities:

$$
\text{Var}(\hat{\theta}^{\text{PM}}) \;<\; \text{Var}(\theta).
$$

This *under-dispersion* is a feature when the goal is individual-level
accuracy (Goal 1), because it reduces mean squared error.
But it is a liability for Goals 2 and 3: the empirical distribution of PM
estimates is too narrow, so tail probabilities are systematically
mis-estimated and extreme ranks are compressed toward the centre.

### Visualising shrinkage

The figure below loads a pre-computed bimodal simulation ($N = 200$,
25 items, Rasch).  The true density has two modes; PM fills the valley.

```{r shrinkage-figure, eval=has_data, fig.height=5, fig.cap="PM (red) compresses the distribution. CB (blue) and GR (green) progressively de-shrink the estimates, with GR tracking the true density most closely."}
comp <- readRDS(find_extdata("vignette_estimates_comparison.rds"))

N <- length(comp$true_theta)
df_shrink <- data.frame(
  value = c(comp$true_theta,
            comp$dpm$theta$theta_pm,
            comp$dpm$theta$theta_cb,
            comp$dpm$theta$theta_gr),
  Method = factor(rep(c("True", "PM", "CB", "GR"), each = N),
                  levels = c("True", "PM", "CB", "GR"))
)

ggplot(df_shrink, aes(x = value, colour = Method, fill = Method)) +
  geom_density(alpha = 0.15, linewidth = 0.9) +
  scale_colour_manual(values = c(True = unname(palette_dpmirt["true"]),
                                  PM   = unname(palette_dpmirt["pm"]),
                                  CB   = unname(palette_dpmirt["cb"]),
                                  GR   = unname(palette_dpmirt["gr"]))) +
  scale_fill_manual(values = c(True = unname(palette_dpmirt["true"]),
                                PM   = unname(palette_dpmirt["pm"]),
                                CB   = unname(palette_dpmirt["cb"]),
                                GR   = unname(palette_dpmirt["gr"]))) +
  labs(title = "Shrinkage and de-shrinkage under a DPM prior",
       subtitle = "Bimodal population, 25 items, 200 persons",
       x = expression(theta), y = "Density") +
  theme_bw() +
  theme(legend.position = "top")
```

> **What to look for.** PM (red) is the most compressed: the valley between
> the two modes is partially filled and the tails are thin.  CB (blue)
> widens the distribution by matching the first two moments, restoring much
> of the true spread.  GR (green) tracks the true density (gray) most
> closely, preserving both modes and the valley between them.

If the pre-computed data are not installed, the figure above will not
render.  See `vignette("simulation-study")` for generation instructions.


## 4. Posterior Mean (PM) Estimator

### Definition

The posterior mean for person $p$ is simply the average of the $S$
retained MCMC draws:

$$
\hat{\theta}_p^{\text{PM}}
  \;=\; \frac{1}{S}\sum_{s=1}^{S}\theta_p^{(s)}.
$$

Under squared-error loss this is the Bayes-optimal point estimate.
DPMirt returns it as `theta_pm` whenever you call `dpmirt_estimates()`.

### Code

```{r pm-code, eval=FALSE}
# Compute only the posterior mean
est <- dpmirt_estimates(fit, methods = "pm")

# Inspect first few values
head(est$theta[, c("theta_pm", "theta_psd")])
```

### When to use PM

| Situation | Recommendation |
|:----------|:---------------|
| Reporting a single ability score for an individual | PM is optimal |
| Building a profile across multiple sub-tests | PM for each sub-test |
| Feeding estimates into a downstream regression | PM minimises attenuation |
| Identifying extreme students for intervention | Prefer GR (see Section 6) |

PM is the safest default when the consumer of the scores will interpret
them one person at a time.


## 5. Constrained Bayes (CB) Estimator

### Idea

Ghosh (1992) proposed a simple moment-matching correction: rescale the
PM estimates so that their first two moments match those of the marginal
empirical distribution.
This removes the systematic under-dispersion caused by shrinkage while
retaining the Bayesian ranking.

### Formula

Let $\bar\eta = \frac{1}{N}\sum_p \hat{\theta}_p^{\text{PM}}$ be the
grand mean of the PMs, $V_\eta = \text{Var}(\hat{\theta}^{\text{PM}})$
their variance, and
$\bar\lambda = \frac{1}{N}\sum_p \text{Var}(\theta_p \mid \text{data})$
the average posterior variance.  Then

$$
\hat{\theta}_p^{\text{CB}}
  \;=\; \bar\eta
  \;+\; (\hat{\theta}_p^{\text{PM}} - \bar\eta)
        \sqrt{1 + \frac{\bar\lambda}{V_\eta}}.
$$

The two constraints are:

1. **Mean preservation:** $\frac{1}{N}\sum_p \hat{\theta}_p^{\text{CB}}
   = \bar\eta$ (same grand mean as PM).
2. **Variance inflation:** $\text{Var}(\hat{\theta}^{\text{CB}})
   = V_\eta + \bar\lambda$ (restores the marginal variance).

The correction factor $\sqrt{1 + \bar\lambda / V_\eta}$ is always
$\geq 1$: CB estimates are always at least as dispersed as PM estimates.

### Code

```{r cb-code, eval=FALSE}
# Compute PM and CB
est <- dpmirt_estimates(fit, methods = c("pm", "cb"))

# CB estimates alongside PM
head(est$theta[, c("theta_pm", "theta_cb")])
```

### Quality flags

When the posterior means have near-zero variance (e.g., a very
short test or extremely strong prior), the CB scaling factor can
become numerically unstable.
DPMirt detects this and falls back to PM:
`est$quality_flags$cb_fallback` will be `TRUE`.
A scaling factor above 5 triggers a warning stored in
`est$quality_flags$cb_extreme_factor`.

```{r cb-flags, eval=FALSE}
# Check quality flags
est$quality_flags$cb_fallback
est$quality_flags$cb_extreme_factor
```

### Limitations

CB is derived under the assumption of an exchangeable Gaussian prior.
When the true latent distribution is strongly multimodal, rescaling the
first two moments alone cannot recover the shape.
For multimodal distributions, GR generally outperforms CB on both ranking
and EDF estimation.


## 6. Triple-Goal (GR) Estimator

### Idea

Shen & Louis (1998) designed an estimator that simultaneously
optimises three properties: it preserves Bayesian rankings,
matches the empirical distribution function, and remains close
to the posterior means.  In DPMirt the estimator is labeled "GR"
after its two key ingredients: the estimated **G** function (the
EDF of the latent trait) and the posterior mean **R**anks.

### Algorithm (four steps)

The full triple-goal algorithm proceeds as follows:

**Step 1. Posterior mean ranks.**
For each MCMC iteration $s$, rank all $N$ persons by their $\theta_p^{(s)}$
values.
Average across iterations to obtain the posterior mean rank:

$$
\bar{R}_p \;=\; \frac{1}{S}\sum_{s=1}^{S} R_p^{(s)}
  \;=\; \sum_{q=1}^{N} P(\theta_q \le \theta_p \mid \text{data}).
$$

**Step 2. Integer ranks.**
Convert fractional posterior mean ranks to integer ranks by ranking them:

$$
\hat{R}_p \;=\; \text{rank}(\bar{R}_p).
$$

**Step 3. ISEL EDF.**
The integrated squared error loss (ISEL) estimator of the latent EDF pools
all posterior samples:

$$
\hat{G}_N(t) \;=\; \frac{1}{N}\sum_{p=1}^{N}
  P(\theta_p \le t \mid \text{data}).
$$

In practice this is the ECDF of the $N \times S$ pooled draws.

**Step 4. Quantile placement.**
Each person's GR estimate is the quantile of $\hat{G}_N$ that corresponds
to their integer rank:

$$
\hat{\theta}_p^{\text{GR}} \;=\; \hat{G}_N^{-1}
  \!\left(\frac{2\hat{R}_p - 1}{2N}\right).
$$

### Intuition

The GR estimator places one estimate at each of $N$ equally spaced
quantiles of the estimated latent distribution.
The ranking step determines *which* quantile each person occupies;
the quantile step determines the *location* of that estimate on the
$\theta$ scale.
Because the quantile grid spans the full estimated distribution,
the GR estimates reproduce its shape --- modes, tails, and gaps
are all represented faithfully.

### Code

```{r gr-code, eval=FALSE}
# Compute all three estimators
est <- dpmirt_estimates(fit, methods = c("pm", "cb", "gr"))

# GR estimates with posterior mean ranks
head(est$theta[, c("theta_pm", "theta_gr", "rbar", "rhat")])
```

### Quality flags: ties

DPMirt breaks ties in posterior mean ranks randomly by default.
Pass `stop_if_ties = TRUE` to raise an error instead:

```{r gr-ties, eval=FALSE}
est <- dpmirt_estimates(fit, methods = "gr", stop_if_ties = TRUE)
```

Ties are unusual with continuous posteriors; when they occur it
typically signals very low test information for a subset of persons.


## 7. Side-by-Side Comparison

This section loads the same pre-computed estimates object and produces
four complementary views.

### 7.1 Density overlay

```{r density-overlay, eval=has_data, fig.height=5}
# Build long-format data for ggplot
df_dens <- data.frame(
  value = c(comp$true_theta,
            comp$dpm$theta$theta_pm,
            comp$dpm$theta$theta_cb,
            comp$dpm$theta$theta_gr),
  method = rep(c("True", "PM", "CB", "GR"),
               each = length(comp$true_theta))
)
df_dens$method <- factor(df_dens$method,
                          levels = c("True", "PM", "CB", "GR"))

ggplot(df_dens, aes(x = value, colour = method, fill = method)) +
  geom_density(alpha = 0.12, linewidth = 0.9) +
  scale_colour_manual(values = c(True = palette_dpmirt["true"],
                                  PM   = palette_dpmirt["pm"],
                                  CB   = palette_dpmirt["cb"],
                                  GR   = palette_dpmirt["gr"])) +
  scale_fill_manual(values = c(True = palette_dpmirt["true"],
                                PM   = palette_dpmirt["pm"],
                                CB   = palette_dpmirt["cb"],
                                GR   = palette_dpmirt["gr"])) +
  labs(title = "True vs. PM vs. CB vs. GR densities",
       x = expression(theta), y = "Density",
       colour = "Method", fill = "Method") +
  theme_bw() +
  theme(legend.position = "top")
```

> **What to look for.** PM (red) is the most compressed.
> CB (blue) widens the distribution but may not recover multimodal
> shape. GR (green) tracks the true density (grey) most closely,
> reproducing both modes and the valley between them.

### 7.2 Caterpillar plot

When a `dpmirt_estimates` object is available, the built-in `plot()`
method produces a caterpillar plot with all three estimators overlaid.

```{r caterpillar, eval=FALSE}
# Built-in caterpillar plot
plot(est, type = "estimates", param = "theta")
```

Persons are sorted by PM.  CB and GR deviate most at the tails.

### 7.3 Rank scatter

```{r rank-scatter, eval=has_data, fig.height=5}
df_rank <- data.frame(
  rank_pm = rank(comp$dpm$theta$theta_pm),
  rank_gr = rank(comp$dpm$theta$theta_gr)
)

ggplot(df_rank, aes(x = rank_pm, y = rank_gr)) +
  geom_point(alpha = 0.5, size = 1.5, colour = palette_dpmirt["dpm"]) +
  geom_abline(intercept = 0, slope = 1, linetype = "dashed",
              colour = "grey50") +
  labs(title = "Rank agreement: PM vs. GR",
       x = "Rank (PM)", y = "Rank (GR)") +
  theme_bw()
```

Rank agreement is typically very high in the interior.  Discrepancies
concentrate in the tails --- precisely the students targeted by
classification decisions.

### 7.4 Table of first 10 persons

```{r table-first10, eval=has_data}
tab10 <- data.frame(
  Person   = 1:10,
  True     = round(comp$true_theta[1:10], 3),
  PM       = round(comp$dpm$theta$theta_pm[1:10], 3),
  CB       = round(comp$dpm$theta$theta_cb[1:10], 3),
  GR       = round(comp$dpm$theta$theta_gr[1:10], 3),
  PM_rank  = rank(-comp$dpm$theta$theta_pm)[1:10],
  GR_rank  = rank(-comp$dpm$theta$theta_gr)[1:10]
)
knitr::kable(tab10, caption = "Estimates and ranks for the first 10 persons.")
```


## 8. The Role of Test Reliability

### Reliability as the dominant moderator

The simulation studies in Lee & Wind show that average test reliability
$\bar{w}$ --- not sample size $N$ --- is the dominant moderator of
estimator performance.
Reliability determines how much shrinkage occurs and therefore how much
room there is for estimator choice (and prior choice) to make a
difference.

The average shrinkage weight $\bar{w}$ is related to the classical
marginal reliability $\rho$ through
$\bar{w} \approx \rho$ for long tests under a Rasch model.
A rough mapping from reliability to test length is:

```{r reliability-table, echo=FALSE}
rel_df <- data.frame(
  Reliability = c("$\\bar{w} \\approx 0.50$",
                   "$\\bar{w} \\approx 0.70$",
                   "$\\bar{w} \\approx 0.90$"),
  Items = c("~10 items", "~23 items", "~61 items"),
  Shrinkage = c("Severe", "Moderate", "Mild"),
  Choice = c(
    "Estimator choice matters MORE than prior choice",
    "Estimator and prior both matter",
    "Prior choice matters MORE for non-normal G"
  )
)
knitr::kable(
  rel_df,
  col.names = c("Reliability $\\bar{w}$", "~Items needed",
                "Shrinkage severity", "What matters most?"),
  caption = paste("Reliability determines whether estimator choice or",
                  "prior choice has more influence on recovery.")
)
```

### Practical implications

- **Low reliability** ($\bar{w} \approx 0.50$): Shrinkage dominates.
  CB provides the largest EDF gains; model flexibility matters less
  because posteriors concentrate around the prior regardless.
- **Moderate reliability** ($\bar{w} \approx 0.70$): Both estimator and
  prior contribute meaningfully.  DPM + GR starts to outperform
  Normal + GR for non-normal populations.  Most screening instruments
  (20--30 items) fall in this regime.
- **High reliability** ($\bar{w} \approx 0.90$): All three estimators
  converge.  Prior choice (DPM vs. Normal) becomes the primary driver
  of EDF and ranking accuracy for non-normal distributions.

> **Rule of thumb.** If you have a short test and a non-normal population,
> start with CB. If you have a long test and suspect non-normality,
> invest in the DPM prior and use GR.

### Visualizing the reliability effect

The figure below makes the abstract table concrete.  Both panels show a
bimodal population estimated with a DPM prior, but the left panel uses a
10-item test ($\bar{w} \approx 0.5$) while the right uses a 25-item test
($\bar{w} \approx 0.8$).

```{r reliability-comparison, eval=(has_data && has_lowrel_data), fig.width=10, fig.height=5, fig.cap="Effect of reliability on estimator behavior.  Left ($\\rho \\approx 0.5$, 10 items): PM (red) collapses the bimodal distribution into a sharp unimodal peak --- the two modes are completely lost.  CB (blue) and GR (green) resist this compression, preserving the two-group structure even with severe shrinkage.  Right ($\\rho \\approx 0.8$, 25 items): all three estimators begin to recover the true bimodal shape, with GR tracking the truth most closely."}
# Load both reliability conditions
comp_hi  <- readRDS(find_extdata("vignette_estimates_comparison.rds"))
comp_lo  <- readRDS(find_extdata("vignette_estimates_comparison_lowrel.rds"))

build_panel <- function(comp_obj, label) {
  N <- length(comp_obj$true_theta)
  data.frame(
    value = c(comp_obj$true_theta,
              comp_obj$dpm$theta$theta_pm,
              comp_obj$dpm$theta$theta_cb,
              comp_obj$dpm$theta$theta_gr),
    Method = factor(rep(c("True", "PM", "CB", "GR"), each = N),
                    levels = c("True", "PM", "CB", "GR")),
    Reliability = label
  )
}

df_rel <- rbind(
  build_panel(comp_lo, "\u03C1 \u2248 0.5  (10 items)"),
  build_panel(comp_hi, "\u03C1 \u2248 0.8  (25 items)")
)
df_rel$Reliability <- factor(df_rel$Reliability,
  levels = c("\u03C1 \u2248 0.5  (10 items)",
             "\u03C1 \u2248 0.8  (25 items)"))

ggplot(df_rel, aes(x = value, colour = Method, fill = Method)) +
  geom_density(alpha = 0.15, linewidth = 0.9) +
  facet_wrap(~ Reliability, ncol = 2) +
  scale_colour_manual(values = c(True = unname(palette_dpmirt["true"]),
                                  PM   = unname(palette_dpmirt["pm"]),
                                  CB   = unname(palette_dpmirt["cb"]),
                                  GR   = unname(palette_dpmirt["gr"]))) +
  scale_fill_manual(values = c(True = unname(palette_dpmirt["true"]),
                                PM   = unname(palette_dpmirt["pm"]),
                                CB   = unname(palette_dpmirt["cb"]),
                                GR   = unname(palette_dpmirt["gr"]))) +
  labs(title = "Reliability determines when estimator choice matters",
       x = expression(theta), y = "Density") +
  theme_bw() +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold", size = 11))
```

At low reliability (left panel), the posterior for each person is heavily
shrunk toward the prior center, so the PM distribution collapses to a
narrow unimodal peak.  CB and GR resist this compression by
inflating the variance (CB) or placing estimates at quantiles of the
estimated EDF (GR), preserving the two-group structure.

At high reliability (right panel), the data speak loudly enough that
even PM preserves the bimodal shape.  The three estimators converge, and
the primary gains come from using the DPM prior rather than the Normal
prior (see Section 5.4 of `vignette("models-and-workflow")`).

### Quantifying the shrinkage

The table below makes the shrinkage problem concrete with numbers,
comparing the standard deviation of each estimator to the true SD.

```{r shrinkage-table, eval=(has_data && has_lowrel_data), echo=FALSE}
# Quantify dispersion for both reliability levels
shrink_stats <- function(comp_obj, rho_label) {
  data.frame(
    Reliability = rho_label,
    Statistic   = c("SD(True)", "SD(PM)", "SD(CB)", "SD(GR)",
                     "Ratio PM/True", "Ratio CB/True", "Ratio GR/True"),
    Value       = c(
      sd(comp_obj$true_theta),
      sd(comp_obj$dpm$theta$theta_pm),
      sd(comp_obj$dpm$theta$theta_cb),
      sd(comp_obj$dpm$theta$theta_gr),
      sd(comp_obj$dpm$theta$theta_pm) / sd(comp_obj$true_theta),
      sd(comp_obj$dpm$theta$theta_cb) / sd(comp_obj$true_theta),
      sd(comp_obj$dpm$theta$theta_gr) / sd(comp_obj$true_theta)
    )
  )
}

shrink_both <- rbind(
  shrink_stats(comp_lo, "\u03C1 \u2248 0.5"),
  shrink_stats(comp_hi, "\u03C1 \u2248 0.8")
)
shrink_both$Value <- round(shrink_both$Value, 3)

# Reshape to wide format for readability
shrink_wide <- reshape(shrink_both,
                        idvar = "Statistic",
                        timevar = "Reliability",
                        direction = "wide")
names(shrink_wide) <- gsub("Value\\.", "", names(shrink_wide))
knitr::kable(shrink_wide,
             row.names = FALSE,
             caption = paste("Dispersion of estimates relative to truth.",
                            "At low reliability, PM retains only ~50% of the",
                            "true SD; CB and GR recover ~80-95%."))
```

> **Reading the table.** The "Ratio" rows reveal the shrinkage severity.
> At $\rho \approx 0.5$, PM compresses the distribution to roughly half
> its true spread (ratio $\approx 0.50$).  CB and GR partially resist
> this compression, recovering about 70% of the true SD
> (ratio $\approx 0.70$).  At $\rho \approx 0.8$, all three estimators
> converge toward --- or even slightly exceed --- the true dispersion
> (ratios near 1.0), confirming that reliability is the dominant
> moderator of shrinkage severity.


## 9. Practical Recommendations

The table below summarises estimator-prior pairings for common
operational goals.

```{r recommendations-table, echo=FALSE}
rec_df <- data.frame(
  Goal = c(
    "Individual scoring",
    "Identify extreme students",
    "Proportion below cutoff",
    "Recover full distribution"
  ),
  Estimator = c("PM", "GR", "GR", "GR"),
  Prior = c(
    "Either (Normal or DPM)",
    "DPM if reliability > 0.7 and non-normal suspected",
    "DPM if reliability > 0.7",
    "DPM strongly recommended"
  )
)
knitr::kable(
  rec_df,
  col.names = c("Primary Goal", "Estimator", "Prior Recommendation"),
  caption = "Quick-reference guide: estimator and prior by goal."
)
```

### Decision workflow

A simple three-step workflow:
1. **Determine your primary goal.** What will the scores be used for?
   If individual feedback, use PM.  If classification, ranking, or
   distributional summaries, use GR.
2. **Assess reliability.** Estimate $\bar{w}$ from KR-20 or the model's
   average posterior variance.  If $\bar{w} < 0.65$, estimator choice
   dominates; if $\bar{w} > 0.80$, prior choice dominates.
3. **Choose the prior.** If reliability is high and there is reason to
   doubt normality (e.g., known floor/ceiling effects, gifted
   subpopulations, mixed curricula), use a DPM prior.  Otherwise a
   Normal prior is sufficient.


## 10. Loss Functions for Evaluation

When true parameter values are available (e.g., in a simulation study),
`dpmirt_loss()` computes the three loss functions that correspond to the
three goals.

### Built-in loss functions

```{r loss-builtin, eval=FALSE}
# Assume 'fit' was run on simulated data with known true values
est  <- dpmirt_estimates(fit, methods = c("pm", "cb", "gr"))
loss <- dpmirt_loss(est, true_theta = sim$theta, true_beta = sim$beta)
print(loss)
```

```{r loss-table-real, eval=has_loss_data, echo=FALSE}
loss_results <- readRDS(find_extdata("vignette_loss_results.rds"))

# DPM-prior losses on bimodal data (rho ~0.8)
loss_dpm <- loss_results$dpm
loss_theta <- loss_dpm[loss_dpm$parameter == "theta",
                       c("method", "msel", "mselr", "ks")]
loss_theta$method <- toupper(loss_theta$method)
knitr::kable(
  loss_theta,
  col.names = c("Method", "MSEL", "MSELR", "KS"),
  digits = c(0, 4, 6, 4),
  row.names = FALSE,
  caption = paste("Loss function results (DPM prior, bimodal population,",
                  "25 items, rho=0.8).  PM wins on MSEL; GR wins on MSELR and KS.")
)
```

The pattern is typical:

- **MSEL:** PM $<$ GR $<$ CB --- posterior mean minimises individual MSE
  by construction.
- **MSELR:** GR $<$ PM $\approx$ CB --- the triple-goal algorithm directly
  optimises ranking loss.
- **KS:** GR $<$ CB $<$ PM --- GR and CB both reduce distributional bias,
  with GR achieving the tightest fit.

When reliability drops, the loss differences across estimators become more
pronounced:

```{r loss-table-lowrel, eval=(has_loss_data && has_lowrel_data), echo=FALSE}
loss_lo <- readRDS(find_extdata("vignette_loss_results_lowrel.rds"))
loss_hi <- readRDS(find_extdata("vignette_loss_results.rds"))

format_loss <- function(loss_obj, rho_label) {
  ld <- loss_obj$dpm[loss_obj$dpm$parameter == "theta",
                     c("method", "msel", "mselr", "ks")]
  ld$Reliability <- rho_label
  ld$method <- toupper(ld$method)
  ld
}

loss_both <- rbind(
  format_loss(loss_lo, "\u03C1 \u2248 0.5"),
  format_loss(loss_hi, "\u03C1 \u2248 0.8")
)

knitr::kable(
  loss_both[, c("Reliability", "method", "msel", "mselr", "ks")],
  col.names = c("Reliability", "Method", "MSEL", "MSELR", "KS"),
  digits = c(0, 0, 4, 6, 4),
  row.names = FALSE,
  caption = paste("Loss comparison across reliability levels.",
                  "At low reliability, estimator differences are amplified",
                  "\u2014 especially for KS, where GR achieves a much tighter",
                  "fit than PM.")
)
```

### Custom loss functions

Supply an arbitrary loss function via `custom_loss`.  The function must
accept two numeric vectors and return a scalar.

```{r loss-custom, eval=FALSE}
# Mean absolute error
mae_loss <- function(est, true) mean(abs(est - true))

loss_mae <- dpmirt_loss(
  est,
  true_theta  = sim$theta,
  true_beta   = sim$beta,
  custom_loss = mae_loss
)
print(loss_mae)
```

See `vignette("simulation-study")` for a full factorial comparison of
loss across reliability levels, sample sizes, latent shapes, and priors.


## 11. Summary and Next Steps

This vignette has shown that:

- The posterior mean (PM) is optimal when the inferential goal is
  individual-level accuracy (Goal 1: minimise MSEL).
- The constrained Bayes (CB) estimator corrects the under-dispersion of
  PM by matching the first two moments of the marginal distribution.
  It is most helpful when reliability is low and a simple correction
  suffices.
- The triple-goal (GR) estimator simultaneously optimises ranking and
  distribution recovery (Goals 2 and 3).  It is the recommended
  estimator whenever classification, selection, or distributional
  summaries are the primary output.
- Test reliability $\bar{w}$, not sample size, is the dominant moderator
  of estimator performance.  Short tests benefit most from estimator
  choice; long tests benefit most from prior choice (DPM vs. Normal).

### What next?

- **Simulation study.** See `vignette("simulation-study")` for a full
  factorial evaluation of prior $\times$ estimator $\times$ reliability
  $\times$ latent shape.
- **Prior elicitation.** See `vignette("prior-elicitation")` for
  principled selection of the DPM concentration parameter $\alpha$
  using the DPprior package.
- **Getting started.** If you have not yet fit your first model, see
  `vignette("getting-started")` for the basic DPMirt workflow.

### References

- Ghosh, M. (1992). Constrained Bayes estimation with applications.
  *Journal of the American Statistical Association, 87*(418), 533--540.
- Lee, J. & Wind, S. Targeting toward inferential goals in Bayesian Rasch
  models for estimating person-specific latent traits. *OSF Preprint*.
  <https://doi.org/10.31219/osf.io/qrw4n>
- Louis, T. A. (1984). Estimating a population of parameter values using
  Bayes and empirical Bayes methods. *Journal of the American Statistical
  Association, 79*(386), 393--398.
- Paganin, S., Paciorek, C. J., Wehrhahn, C., Rodríguez, A.,
  Rabe-Hesketh, S., & de Valpine, P. (2023). Computational strategies and
  estimation performance with Bayesian semiparametric item response theory
  models. *Journal of Educational and Behavioral Statistics, 48*(2),
  147--188. https://doi.org/10.3102/10769986221136105
- Shen, W., & Louis, T. A. (1998). Triple-goal estimates in two-stage
  hierarchical models. *Journal of the Royal Statistical Society: Series B,
  60*(2), 455--471.

```{r session-info, echo=FALSE}
sessionInfo()
```
