---
title: "Principled Prior Elicitation for the DP Concentration Parameter"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  Learn why the DP concentration parameter alpha matters, how the Paganin

  default compares to principled elicitation via DPprior, and how to integrate
  both approaches into your DPMirt workflow --- including sensitivity analysis
  and custom base measure tuning.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Principled Prior Elicitation for the DP Concentration Parameter}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(DPMirt)
knitr::opts_chunk$set(
  collapse    = TRUE,
  comment     = "#>",
  fig.width   = 7,
  fig.height  = 5,
  fig.align   = "center",
  out.width   = "85%"
)
set.seed(42)

# DPMirt palette
pal <- list(
  parametric     = "#56B4E9",
  semiparametric = "#E69F00",
  reference      = "gray50"
)

# Helper: find extdata files in installed package OR source tree
find_extdata <- function(filename) {
  path <- system.file("extdata", filename, package = "DPMirt")
  if (nzchar(path) && file.exists(path)) return(path)
  path2 <- file.path("..", "inst", "extdata", filename)
  if (file.exists(path2)) return(path2)
  ""
}

# Guard: pre-computed sensitivity results must exist for eval'd chunks
has_data <- nzchar(find_extdata("vignette_sensitivity_fits.rds"))

# Guard: DPprior available for live elicitation calls
has_dpprior <- requireNamespace("DPprior", quietly = TRUE)
```

## Overview

Every Dirichlet Process Mixture (DPM) model contains a concentration parameter
$\alpha$ that controls how much the nonparametric prior deviates from a single
parametric component.
Choosing $\alpha$ well is not just a technical detail --- it shapes cluster
formation, posterior shrinkage, and the balance between flexibility and
parsimony.

This vignette covers:

1. **Why $\alpha$ matters** --- its three operational dimensions.
2. **Paganin default vs. principled elicitation** --- when $\text{Gamma}(1,3)$
   is adequate and when it is not.
3. **DPprior integration** --- three routes to set $\alpha$ in DPMirt.
4. **Two strategies from Lee (2026)** --- DP-diffuse and DP-inform.
5. **Sensitivity analysis** --- comparing posteriors under different priors.
6. **Fallback behavior** --- what happens when DPprior is not installed.
7. **Custom base measure tuning** --- advanced hyperparameter control.

## Why Alpha Matters

The concentration parameter $\alpha$ operates along three dimensions that
jointly determine the behavior of the DP prior.

### Dimension 1: Number of Clusters

The expected number of distinct clusters $K$ among $N$ observations is
approximately

$$
\mathrm{E}[K \mid \alpha, N] \approx \alpha \log\!\left(1 + \frac{N}{\alpha}\right).
$$

Larger $\alpha$ implies more clusters; smaller $\alpha$ concentrates the
posterior onto fewer groups.

### Dimension 2: Weight Concentration

The first stick-breaking weight $w_1$ follows $\text{Beta}(1, \alpha)$.
When $\alpha$ is small, $\mathrm{E}[w_1]$ is large and most mass is
captured by a single component.
When $\alpha$ is large, mass spreads across many components.

### Dimension 3: Posterior Shrinkage

In IRT, the prior on $\theta$ acts as a regularizer.
A DPM prior with small $\alpha$ shrinks person abilities toward fewer
cluster centers, closely resembling a parametric Normal prior.
A DPM prior with large $\alpha$ allows each person to retain more of their
data-driven estimate, reducing shrinkage at the cost of higher variance.

> **Key insight:** The choice of $\alpha$ is not merely a tuning knob ---
> it encodes a substantive belief about the heterogeneity of the latent
> trait distribution.  Getting it wrong can either mask genuine subgroups
> (too small) or fragment a homogeneous population (too large).

## Paganin Default vs. Principled Elicitation

Paganin et al. (2023) use $\alpha \sim \text{Gamma}(1, 3)$ as a default.
This prior has mean $\mathrm{E}[\alpha] = 1/3$ and is concentrated near
zero, strongly favoring few clusters.

For many educational testing applications where the latent trait is
approximately unimodal, this default is reasonable.
However, it can be overly restrictive when:
- The population is known to be heterogeneous (e.g., mixed clinical groups).
- Sample size is large ($N > 500$) and the data can support many clusters.
- The research question specifically concerns subgroup detection.

Lee (2026) provides a principled alternative through the **DPprior** package,
which uses Two-Stage Moment Matching (TSMM) to translate a researcher's belief
about the expected number of clusters into a calibrated
$\text{Gamma}(a, b)$ hyperprior. The updated DPprior package also provides
diagnostic tools for assessing dominance risk and prior quality, as well as
a dual-anchor calibration framework for weight-constrained elicitation.

```{r paganin-default, fig.cap = "The Paganin default Gamma(1, 3) prior for alpha, with implied E[alpha] = 0.33."}
alpha_grid <- seq(0, 5, length.out = 500)
dens_default <- dgamma(alpha_grid, shape = 1, rate = 3)

plot(alpha_grid, dens_default, type = "l", lwd = 2, col = pal$reference,
     xlab = expression(alpha), ylab = "Density",
     main = expression(paste("Paganin default: ", alpha, " ~ Gamma(1, 3)")))
abline(v = 1/3, lty = 2, col = pal$reference)
text(1.2, max(dens_default) * 0.8,
     bquote(E[alpha] == .(round(1/3, 2))), col = pal$reference)
```

## DPprior Integration

DPMirt offers three ways to set the $\alpha$ prior, ranging from fully
automated to fully manual.

### Method 1: `dpmirt_alpha_prior()` Wrapper

The simplest approach is the standalone elicitation function, which returns
a named vector `c(a = ..., b = ...)`:

```{r alpha-prior-wrapper, eval = has_dpprior}
# Expect ~5 clusters among 200 persons, medium confidence
ab <- dpmirt_alpha_prior(N = 200, mu_K = 5, confidence = "medium")
ab
```

```{r alpha-prior-wrapper-show, eval = !has_dpprior}
# DPprior not installed --- showing expected output
cat("Alpha prior: Gamma(1.60, 0.82) [E[alpha]=1.96]\n")
cat("  a     b \n")
cat("1.60  0.82\n")
```

### Method 2: Elicit Then Pass to `dpmirt()`

Elicit the prior first with `dpmirt_alpha_prior()`, then pass the
result directly as `alpha_prior`:

```r
ab <- dpmirt_alpha_prior(N = 200, mu_K = 5, confidence = "medium")

fit <- dpmirt(
  data,
  model       = "rasch",
  prior       = "dpm",
  alpha_prior = ab,
  niter       = 10000,
  nburnin     = 2000
)
```

### Method 3: Manual DPprior Workflow

For maximum control, call DPprior directly and pass the result:

```{r dpprior-manual, eval = has_dpprior}
library(DPprior)

fit_prior <- DPprior_fit(J = 200, mu_K = 5, confidence = "medium")
print(fit_prior)
```

```{r dpprior-manual-use, eval = FALSE}
# Pass the DPprior_fit object to dpmirt
fit <- dpmirt(
  data,
  model       = "rasch",
  prior       = "dpm",
  alpha_prior = fit_prior,
  niter       = 10000
)
```

### Method 4: Return the Full DPprior_fit Object

For access to diagnostics and convergence information, use `return_fit = TRUE`:

```{r alpha-prior-full, eval = has_dpprior}
# Get the full DPprior_fit object
fit_prior <- dpmirt_alpha_prior(N = 200, mu_K = 5, confidence = "medium",
                                 return_fit = TRUE)

# Inspect the result
print(fit_prior)

# Pass directly to dpmirt (DPprior_fit objects are accepted)
# fit <- dpmirt(data, prior = "dpm", alpha_prior = fit_prior)
```

```{r alpha-prior-full-fallback, eval = !has_dpprior, echo = FALSE}
cat("DPprior not installed --- the return_fit option requires DPprior.\n")
```

### Confidence Levels

The `confidence` argument controls how tightly the prior concentrates
around the expected cluster count.
It maps to the variance of the implied prior on $K$:

```{r confidence-table, eval = has_dpprior}
conf_levels <- c("low", "medium", "high")
conf_results <- lapply(conf_levels, function(cl) {
  fp <- DPprior_fit(J = 200, mu_K = 5, confidence = cl)
  data.frame(
    Confidence = cl,
    a          = round(fp$a, 3),
    b          = round(fp$b, 3),
    E_alpha    = round(fp$a / fp$b, 3),
    CV_alpha   = round(1 / sqrt(fp$a), 3),
    Var_K      = round(fp$target$var_K, 2),
    stringsAsFactors = FALSE
  )
})
conf_table <- do.call(rbind, conf_results)
knitr::kable(conf_table,
             caption = "Confidence levels and their implied hyperparameters (N = 200, mu_K = 5).",
             align = "lccccc")
```

```{r confidence-table-fallback, eval = !has_dpprior, echo = FALSE}
# Fallback display when DPprior is not installed
cat("Confidence |   a   |   b   | E[alpha] | CV[alpha] | Var(K)\n")
cat("-----------|-------|-------|----------|-----------|-------\n")
cat("low        | 0.640 | 0.327 |   1.960  |   1.250   | 12.50\n")
cat("medium     | 1.600 | 0.816 |   1.960  |   0.791   |  5.00\n")
cat("high       | 4.000 | 2.041 |   1.960  |   0.500   |  2.00\n")
```

> **Interpretation:** Higher confidence concentrates the Gamma prior
> (lower CV), constraining $\alpha$ and thereby the cluster count.
> Low confidence yields a diffuse prior that lets the data determine
> cluster structure.  All three levels share the same prior mean for
> $\alpha$, differing only in spread.

Note that DPprior calls are analytic (closed-form or fast Newton iteration)
and complete in milliseconds --- no MCMC is involved.

## Two Strategies from Lee (2026)

The APM manuscript (Lee & Wind) identifies two complementary approaches
to specifying $\alpha$:

### DP-Diffuse (Exploratory)

Use when you have little domain knowledge about population heterogeneity.
The goal is a prior that is minimally informative yet properly calibrated
to the sample size:

```{r dp-diffuse, eval = has_dpprior}
# DP-diffuse: set mu_K to log(N), low confidence
N <- 200
mu_K_diffuse <- max(3, ceiling(log(N)))  # ~5 for N=200

ab_diffuse <- dpmirt_alpha_prior(N = N, mu_K = mu_K_diffuse,
                                  confidence = "low")
cat("DP-diffuse: Gamma(", round(ab_diffuse["a"], 2), ", ",
    round(ab_diffuse["b"], 2), ")\n", sep = "")
cat("  E[alpha] =", round(ab_diffuse["a"] / ab_diffuse["b"], 2), "\n")
```

This is the strategy DPMirt uses when `mu_K = NULL` is left at its
default.

### DP-Inform (Domain-Driven)

Use when substantive knowledge suggests a specific number of latent
subgroups.
For example, an assessment designed to distinguish remedial, proficient,
and advanced learners might target $\mu_K = 3$:

```{r dp-inform, eval = has_dpprior}
# DP-inform: domain-driven belief of 3 subgroups, high confidence
ab_inform <- dpmirt_alpha_prior(N = 200, mu_K = 3, confidence = "high")
cat("DP-inform: Gamma(", round(ab_inform["a"], 2), ", ",
    round(ab_inform["b"], 2), ")\n", sep = "")
cat("  E[alpha] =", round(ab_inform["a"] / ab_inform["b"], 2), "\n")
```

### Visual Comparison

```{r compare-strategies, fig.cap = "Comparing DP-diffuse and DP-inform alpha priors.", eval = has_dpprior, fig.height = 4}
alpha_grid <- seq(0.01, 8, length.out = 500)

dens_diffuse <- dgamma(alpha_grid, shape = ab_diffuse["a"],
                        rate = ab_diffuse["b"])
dens_inform  <- dgamma(alpha_grid, shape = ab_inform["a"],
                        rate = ab_inform["b"])
dens_paganin <- dgamma(alpha_grid, shape = 1, rate = 3)

plot(alpha_grid, dens_diffuse, type = "l", lwd = 2, col = pal$semiparametric,
     xlab = expression(alpha), ylab = "Density",
     ylim = c(0, max(c(dens_diffuse, dens_inform, dens_paganin)) * 1.1),
     main = "Alpha prior comparison")
lines(alpha_grid, dens_inform, lwd = 2, col = pal$parametric)
lines(alpha_grid, dens_paganin, lwd = 2, col = pal$reference, lty = 2)
legend("topright",
       legend = c("DP-diffuse (low, mu_K=5)",
                   "DP-inform (high, mu_K=3)",
                   "Paganin Gamma(1,3)"),
       col    = c(pal$semiparametric, pal$parametric, pal$reference),
       lwd    = 2, lty = c(1, 1, 2), bty = "n")
```

## Sensitivity Analysis

A responsible Bayesian analysis examines how conclusions change under
different prior specifications.
Here we compare posterior results from three $\alpha$ priors applied to
the same data.

The fits below were pre-computed.
Each used a Rasch-DPM model on the same bimodal data set ($N = 200$,
$I = 25$):

```{r load-sensitivity, eval = has_data}
sens <- readRDS(find_extdata("vignette_sensitivity_fits.rds"))

# sens contains:
#   $default   - Gamma(1, 3) [Paganin default]
#   $broad     - DP-diffuse
#   $inform    - DP-inform
```

### Posterior Density Comparison

```{r sensitivity-density, eval = has_data, fig.cap = "Posterior mean theta densities under three alpha priors.  When the data are informative, the posteriors converge despite different priors."}
# Extract posterior mean theta for each fit
est_paganin <- dpmirt_estimates(sens$default, methods = "pm")
est_diffuse <- dpmirt_estimates(sens$broad, methods = "pm")
est_inform  <- dpmirt_estimates(sens$inform,  methods = "pm")

dens_p <- density(est_paganin$theta$theta_pm)
dens_d <- density(est_diffuse$theta$theta_pm)
dens_i <- density(est_inform$theta$theta_pm)

xlim <- range(c(dens_p$x, dens_d$x, dens_i$x))
ylim <- c(0, max(c(dens_p$y, dens_d$y, dens_i$y)) * 1.1)

plot(dens_p, xlim = xlim, ylim = ylim, col = pal$reference, lwd = 2,
     lty = 2, main = "Posterior theta density under different alpha priors",
     xlab = expression(theta), ylab = "Density")
lines(dens_d, col = pal$semiparametric, lwd = 2)
lines(dens_i, col = pal$parametric, lwd = 2)
legend("topright",
       legend = c("Paganin Gamma(1,3)", "DP-diffuse", "DP-inform"),
       col = c(pal$reference, pal$semiparametric, pal$parametric),
       lwd = 2, lty = c(2, 1, 1), bty = "n")
```

### Cluster Count Comparison

```{r sensitivity-clusters, eval = has_data}
cluster_summary <- data.frame(
  Prior          = c("Paganin Gamma(1,3)", "DP-diffuse", "DP-inform"),
  Median_K       = c(median(sens$default$cluster_info$n_clusters),
                     median(sens$broad$cluster_info$n_clusters),
                     median(sens$inform$cluster_info$n_clusters)),
  Mean_K         = c(mean(sens$default$cluster_info$n_clusters),
                     mean(sens$broad$cluster_info$n_clusters),
                     mean(sens$inform$cluster_info$n_clusters)),
  Posterior_alpha = c(
    mean(sens$default$other_samp[, "alpha"]),
    mean(sens$broad$other_samp[, "alpha"]),
    mean(sens$inform$other_samp[, "alpha"])
  ),
  WAIC           = c(sens$default$waic,
                     sens$broad$waic,
                     sens$inform$waic)
)
cluster_summary$Median_K <- round(cluster_summary$Median_K, 1)
cluster_summary$Mean_K   <- round(cluster_summary$Mean_K, 1)
cluster_summary$Posterior_alpha <- round(cluster_summary$Posterior_alpha, 3)
cluster_summary$WAIC     <- round(cluster_summary$WAIC, 1)

knitr::kable(cluster_summary,
             caption = "Sensitivity analysis: cluster counts and WAIC under three alpha priors.",
             align = "lcccc")
```

> **Takeaway:** When data are moderately informative (say, $\bar{w} \ge 0.7$
> and $N \ge 100$), the posterior distributions of theta tend to be
> robust to the choice of alpha prior.  The main differences appear in
> the number of active clusters and the posterior uncertainty around alpha
> itself.  WAIC can help discriminate, but differences are often small.

The convergence observed here reflects the moderately high reliability
($\bar{w} \approx 0.8$) of the 25-item test.  For shorter tests with
lower reliability ($\bar{w} \approx 0.5$), the alpha prior may exert
stronger influence on the posterior because the data provide less
information to overcome the prior.  See `vignette("posterior-summaries")`
for a detailed analysis of how reliability moderates both estimator
choice and prior choice.

## Fallback Behavior

When the DPprior package is not installed, DPMirt falls back gracefully
to the Paganin default with an informative message:

```{r fallback-demo}
# Simulate what happens without DPprior
# (dpmirt_alpha_prior handles this internally)
if (!has_dpprior) {
  ab_fallback <- dpmirt_alpha_prior(N = 200)
  cat("Returned:", ab_fallback, "\n")
} else {
  cat("With DPprior installed, the package uses principled elicitation.\n")
  cat("Without DPprior, it returns c(a = 1, b = 3) with a message:\n\n")
  cat('  "DPprior not installed. Using default: Gamma(1, 3)\n')
  cat('   [Paganin et al., 2023].\n')
  cat('   Install DPprior for principled alpha elicitation:\n')
  cat('   remotes::install_github(\\"joonho112/DPprior\\")"\n')
}
```

This design ensures that DPMirt works out of the box while encouraging
users to install DPprior for better-calibrated priors.
DPprior is listed in `Suggests`, not `Imports`, so it is never a hard
dependency.

## Advanced: Custom Base Measure

The DP base measure $G_0$ determines the shape of each mixture component.
DPMirt uses a Normal--Inverse-Gamma base:

$$
G_0 = \mathcal{N}(0, s^2_\mu) \times \text{InvGamma}(\nu_1, \nu_2)
$$

The Paganin defaults are $s^2_\mu = 2$, $\nu_1 = 2.01$, $\nu_2 = 1.01$.
These yield wide cluster means and diffuse cluster variances,
appropriate for standardized IRT scales.

To customize:

```{r custom-base-measure, eval = FALSE}
# Wider cluster means, more concentrated cluster variances
fit <- dpmirt(
  data,
  model        = "rasch",
  prior        = "dpm",
  mu_K         = 5,
  confidence   = "medium",
  base_measure = list(s2_mu = 3, nu1 = 2.01, nu2 = 1.01),
  niter        = 10000
)
```

The step-by-step interface gives direct access:

```{r custom-base-spec, eval = FALSE}
spec <- dpmirt_spec(
  data,
  model        = "rasch",
  prior        = "dpm",
  alpha_prior  = c(a = 1.6, b = 0.82),
  base_measure = list(s2_mu = 3, nu1 = 2.01, nu2 = 1.01)
)

# Inspect what was set
spec$constants$s2_mu   # 3
spec$constants$nu1     # 2.01
spec$constants$nu2     # 1.01
```

### Base Measure Hyperparameter Effects

| Hyperparameter | Default | Effect of Increasing |
|:--------------:|:-------:|:---------------------|
| $s^2_\mu$      | 2       | Wider spread of cluster means --- clusters can be further apart |
| $\nu_1$        | 2.01    | Tighter cluster variances (more concentrated components) |
| $\nu_2$        | 1.01    | Wider cluster variances (more diffuse components) |

> **Practical guidance:** For most IRT applications, the Paganin defaults
> work well.  Increase $s^2_\mu$ if you expect latent subgroups separated
> by more than 3--4 standard deviations (e.g., combined clinical and
> non-clinical samples).

## Elicitation Decision Flowchart

The following summary may help decide which approach to use:

| Scenario | Recommended approach | Code |
|:---------|:---------------------|:-----|
| No domain knowledge, exploratory | DP-diffuse | `dpmirt(data, prior = "dpm")` |
| Substantive belief about subgroups | DP-inform | `dpmirt(data, prior = "dpm", mu_K = 3, confidence = "high")` |
| Reproducing Paganin et al. | Manual Gamma(1,3) | `dpmirt(data, prior = "dpm", alpha_prior = c(1, 3))` |
| Comparing priors | Sensitivity analysis | Fit multiple models, compare WAIC and posteriors |
| DPprior not available | Automatic fallback | `dpmirt(data, prior = "dpm")` --- uses Gamma(1,3) with message |

## What's Next?

Now that you can set principled priors for $\alpha$, the following
vignettes show how to put them to work:

| Vignette | Why read it |
|:---------|:------------|
| *Quick Start* | End-to-end pipeline with data simulation, fitting, and estimation |
| *Simulation Study* | Full factorial study comparing Normal vs. DPM priors under different conditions |
| *Under the Hood* | How DPMirt generates NIMBLE code, compiles, and samples |

## References

Lee, J. (2026). Design-conditional prior elicitation for Dirichlet
Process mixtures: A unified framework for cluster counts and weight
control. *arXiv preprint arXiv:2602.06301*.
<https://arxiv.org/abs/2602.06301>

Lee, J. & Wind, S. Targeting toward inferential goals in Bayesian Rasch
models for estimating person-specific latent traits. *OSF Preprint*.
<https://doi.org/10.31219/osf.io/qrw4n>

Paganin, S., Paciorek, C. J., Wehrhahn, C., Rodríguez, A.,
Rabe-Hesketh, S., & de Valpine, P. (2023). Computational strategies
and estimation performance with Bayesian semiparametric item response
theory models. *Journal of Educational and Behavioral Statistics*,
48(2), 147--188. https://doi.org/10.3102/10769986221136105
