---
title: "Simulation Study: Evaluating Prior Models and Posterior Summaries"
author: "JoonHo Lee"
date: "`r Sys.Date()`"
description: >
  A comprehensive guide to designing, executing, and analyzing simulation
  studies with DPMirt --- covering factorial design, IRTsimrel integration,
  the full analysis pipeline, loss function evaluation, and key findings
  from the APM manuscript.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Study: Evaluating Prior Models and Posterior Summaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
library(DPMirt)
library(ggplot2)
knitr::opts_chunk$set(
  collapse    = TRUE,
  comment     = "#>",
  fig.width   = 7,
  fig.height  = 5,
  fig.align   = "center",
  out.width   = "85%"
)
set.seed(42)

# DPMirt palette
pal <- list(
  parametric     = "#56B4E9",
  semiparametric = "#E69F00",
  reference      = "gray50",
  normal_pm      = "#56B4E9",
  normal_cb      = "#0072B2",
  normal_gr      = "#004080",
  dpm_pm         = "#E69F00",
  dpm_cb         = "#D55E00",
  dpm_gr         = "#CC3311"
)

# Helper: find extdata files in installed package OR source tree
find_extdata <- function(filename) {
  path <- system.file("extdata", filename, package = "DPMirt")
  if (nzchar(path) && file.exists(path)) return(path)
  path2 <- file.path("..", "inst", "extdata", filename)
  if (file.exists(path2)) return(path2)
  ""
}

# Guard: pre-computed results must exist
has_data <- nzchar(find_extdata("vignette_sim_bimodal.rds")) &&
  nzchar(find_extdata("vignette_estimates_comparison.rds")) &&
  nzchar(find_extdata("vignette_loss_results.rds"))
has_lowrel_data <- nzchar(find_extdata("vignette_estimates_comparison_lowrel.rds")) &&
  nzchar(find_extdata("vignette_loss_results_lowrel.rds"))
```

## Overview

Simulation studies are the primary tool for evaluating IRT methods because
they provide access to the ground truth --- the true person abilities and
item parameters that generated the data.
DPMirt provides an integrated simulation framework that lets you:

1. **Simulate** data under known conditions using IRTsimrel calibration for
   Rasch/2PL models, or DPMirt's built-in fallback.
2. **Fit** models under competing prior specifications (Normal vs. DPM).
3. **Estimate** person abilities using three posterior summary methods
   (PM, CB, GR).
4. **Evaluate** performance using multiple loss functions (MSEL, MSELR, KS).

This vignette walks through a complete simulation study design, executes
a single-condition deep dive, interprets the results, and provides
templates for scaling to a full factorial study.

## Simulation Design

The APM manuscript (Lee & Wind) uses a $3 \times 2 \times 3$ factorial
design crossing three factors:

### Factor 1: Latent Distribution Shape ($G$)

| Shape | Description | Generating mechanism |
|:------|:------------|:---------------------|
| Normal | Standard Gaussian | $\theta \sim N(0, 1)$ |
| Bimodal | Two-group mixture | $0.5 \cdot N(-1.5, 0.5) + 0.5 \cdot N(1.5, 0.5)$ |
| Skew | Right-skewed | Shifted exponential: $\text{Exp}(1) - 1$ |

The **Normal** condition is the parametric model's "home turf" where
DPM offers little advantage.
The **Bimodal** and **Skew** conditions represent departures where
flexible priors may help, especially for distributional summaries.

### Factor 2: Sample Size ($N$)

| $N$ | Interpretation |
|:----|:---------------|
| 50 | Small sample: high uncertainty, strong shrinkage |
| 200 | Moderate sample: reliable estimation for most methods |

### Factor 3: Marginal Reliability ($\bar{w}$)

Reliability $\bar{w}$ is the most influential factor in the simulation.
It controls the information per person and therefore the degree of
posterior shrinkage.

```{r reliability-table, echo = FALSE}
rel_table <- data.frame(
  `Reliability` = c("0.50", "0.70", "0.90"),
  `Approx_Items` = c("~10", "~23", "~61"),
  Shrinkage = c("Severe", "Moderate", "Mild"),
  Separation = c("1.0", "1.5", "3.0"),
  stringsAsFactors = FALSE
)
names(rel_table) <- c("w-bar", "~Items (Rasch)", "Shrinkage", "Separation")
knitr::kable(rel_table,
             caption = "Reliability levels and their implications for IRT estimation.",
             align = "cccc")
```

> **Why reliability dominates:** At $\bar{w} = 0.50$, each person's
> posterior is heavily shrunk toward the prior, so prior differences are
> harder to see and estimator choice can dominate distributional recovery.
> At $\bar{w} = 0.90$, posteriors are tightly concentrated around
> data-driven estimates, making prior misspecification easier to detect.

### Full Design Matrix

The full factorial design has $3 \times 2 \times 3 = 18$ conditions.
Each condition is replicated (e.g., 100 times) for Monte Carlo stability.

```{r design-matrix, echo = FALSE}
design <- expand.grid(
  G     = c("Normal", "Bimodal", "Skew"),
  N     = c(50, 200),
  w_bar = c(0.50, 0.70, 0.90),
  stringsAsFactors = FALSE
)
design$Condition <- seq_len(nrow(design))
design <- design[, c("Condition", "G", "N", "w_bar")]
knitr::kable(design,
             caption = "Full 3 x 2 x 3 factorial design (18 conditions).",
             align = "clcc")
```

### IRTsimrel Integration

DPMirt uses the **IRTsimrel** package (when available) to achieve precise
reliability targeting for Rasch and 2PL simulations. The default
average-information metric uses Empirical Quadrature Calibration (EQC);
MSEM-based targets use IRTsimrel's SAC calibration.
The key function is `dpmirt_simulate()`:

```{r simulate-example}
# Simulate bimodal data with target reliability 0.8
sim <- dpmirt_simulate(
  n_persons    = 200,
  n_items      = 25,
  model        = "rasch",
  target_rho   = 0.8,
  latent_shape = "bimodal",
  seed         = 42
)

sim
```

The returned `dpmirt_sim` object reports:

- **Method**: "irtsimrel" (IRTsimrel-calibrated) or "fallback" (Paganin-style)
- **KR-20**: Empirical reliability of the generated data
- **Achieved rho**: IRTsimrel calibration-level achieved marginal
  reliability (`sim$achieved_rho`) when IRTsimrel was used; `NULL` for
  fallback simulations
- **Target rho**: the requested reliability when IRTsimrel was used; `NULL`
  for fallback simulations
- **Calibration c***: the calibration constant when IRTsimrel was used

When IRTsimrel is not installed, or when `model = "3pl"`,
`dpmirt_simulate()` falls back to a simpler simulation with evenly spaced
item difficulties and no reliability targeting. 3PL fallback simulations
include Beta-distributed guessing parameters `delta`, and `target_rho` is
not used.

### Visualizing the Simulated Data

```{r sim-theta-density, fig.cap = "Density of true theta values from the bimodal simulation."}
theta_dens <- density(sim$theta)
plot(theta_dens, main = "True theta distribution (bimodal)",
     xlab = expression(theta), lwd = 2)
rug(sim$theta, col = adjustcolor("black", alpha.f = 0.2))
```

```{r sim-sumscores, fig.cap = "Distribution of sum scores from the simulated response matrix."}
sum_scores <- rowSums(sim$response)
hist(sum_scores, breaks = 20, main = "Sum score distribution",
     xlab = "Sum score", col = adjustcolor(pal$parametric, alpha.f = 0.5),
     border = "white")
```

## Analysis Pipeline

The full analysis pipeline for a single simulation condition consists
of six steps.
Below we show the complete code you would run in a live session, then
load pre-computed results for the remaining analysis.

### Step 1: Simulate Data

```{r pipeline-simulate}
sim <- dpmirt_simulate(
  n_persons    = 200,
  n_items      = 25,
  model        = "rasch",
  target_rho   = 0.8,
  latent_shape = "bimodal",
  seed         = 42
)
```

### Step 2: Fit Models

```r
# --- Fit Normal prior model (~2 min compilation + sampling) ---
fit_normal <- dpmirt(
  sim$response,
  model   = "rasch",
  prior   = "normal",
  niter   = 10000,
  nburnin = 2000,
  thin    = 32,
  thin2   = 32,
  seed    = 100
)

# --- Fit DPM prior model (~2 min compilation + sampling) ---
# Bundled fixtures use the conservative Gamma(1,3) alpha prior.
fit_dpm <- dpmirt(
  sim$response,
  model   = "rasch",
  prior   = "dpm",
  niter   = 10000,
  nburnin = 2000,
  thin    = 32,
  thin2   = 32,
  seed    = 101
)
```

### Step 3: Extract Estimates

```r
# Compute PM, CB, GR for both models
est_normal <- dpmirt_estimates(fit_normal, methods = c("pm", "cb", "gr"))
est_dpm    <- dpmirt_estimates(fit_dpm,    methods = c("pm", "cb", "gr"))
```

### Step 4: Evaluate Losses

```r
# Compare to true theta
loss_normal <- dpmirt_loss(est_normal, true_theta = sim$theta,
                            true_beta = sim$beta)
loss_dpm    <- dpmirt_loss(est_dpm,    true_theta = sim$theta,
                            true_beta = sim$beta)
```

### Step 5: Combine Results

```r
loss_normal$prior <- "Normal"
loss_dpm$prior    <- "DPM"
combined_loss <- rbind(loss_normal, loss_dpm)
```

### Loading Pre-Computed Results

For this vignette, all fitting results are pre-computed:
the bundled DPM fixtures use the conservative $\text{Gamma}(1,3)$ prior
for $\alpha$, so the example does not require DPprior.

```{r load-precomputed, eval = has_data}
# Load pre-computed simulation data
sim_data <- readRDS(find_extdata("vignette_sim_bimodal.rds"))

# Load estimates comparison (all 6 method-prior combos)
est_comparison <- readRDS(find_extdata("vignette_estimates_comparison.rds"))

# Load loss results
loss_results <- readRDS(find_extdata("vignette_loss_results.rds"))
```

## Single Condition Deep Dive

We focus on the most informative condition:
**Bimodal**, $N = 200$, $\bar{w} \approx 0.8$.
This condition is useful because the true latent distribution violates the
Gaussian assumption while the data are informative enough to compare
posterior summaries. In the bundled fixture, the DPM prior has its clearest
edge on the KS distributional loss, while Normal-prior GR is slightly better
on the rank loss.

### Posterior Density Overlay

```{r density-overlay, eval = has_data, fig.height = 6, fig.cap = "Density estimates from six method-prior combinations overlaid with the true latent density.  In this fixture, DPM-GR gives the smallest KS loss while posterior means remain visibly shrunk."}
theta_true <- est_comparison$true_theta
N <- length(theta_true)

df_dens <- data.frame(
  value = c(theta_true,
            est_comparison$normal$theta$theta_pm,
            est_comparison$normal$theta$theta_cb,
            est_comparison$normal$theta$theta_gr,
            est_comparison$dpm$theta$theta_pm,
            est_comparison$dpm$theta$theta_cb,
            est_comparison$dpm$theta$theta_gr),
  Method = factor(rep(c("True", "Normal-PM", "Normal-CB", "Normal-GR",
                         "DPM-PM", "DPM-CB", "DPM-GR"), each = N),
                  levels = c("True", "Normal-PM", "Normal-CB", "Normal-GR",
                             "DPM-PM", "DPM-CB", "DPM-GR"))
)

method_colors <- c(
  "True"      = "black",
  "Normal-PM" = pal$normal_pm, "Normal-CB" = pal$normal_cb,
  "Normal-GR" = pal$normal_gr,
  "DPM-PM"    = pal$dpm_pm, "DPM-CB" = pal$dpm_cb,
  "DPM-GR"    = pal$dpm_gr
)
method_lty <- c("True" = "solid",
                "Normal-PM" = "solid", "Normal-CB" = "dashed",
                "Normal-GR" = "dotted",
                "DPM-PM"    = "solid", "DPM-CB"    = "dashed",
                "DPM-GR"    = "dotted")

ggplot(df_dens, aes(x = value, colour = Method, linetype = Method)) +
  geom_density(linewidth = 0.9, fill = NA) +
  scale_colour_manual(values = method_colors) +
  scale_linetype_manual(values = method_lty) +
  labs(title = "Posterior density comparison: 6 methods vs. truth",
       subtitle = "Bimodal population, 25 items, 200 persons",
       x = expression(theta), y = "Density") +
  theme_bw() +
  theme(legend.position = "right")
```

> **What to look for:** Posterior means are visibly compressed under both
> priors. CB and GR de-shrink the estimates, and in this fixture DPM-GR has
> the smallest KS distance to the true distribution. Ranking and individual
> MSE should be read from the loss table rather than inferred from the
> density alone.

### Shrinkage Comparison

```{r shrinkage-plot, eval = has_data, fig.height = 5, fig.cap = "Posterior mean (PM) estimates vs. true theta.  PM estimates are shrunk toward the center under both priors in this fixture."}
par(mfrow = c(1, 2))

# Normal-PM
plot(theta_true, est_comparison$normal$theta$theta_pm,
     pch = 16, cex = 0.6, col = adjustcolor(pal$normal_pm, 0.6),
     xlab = expression(theta[true]), ylab = expression(hat(theta)[PM]),
     main = "Normal-PM", asp = 1)
abline(0, 1, col = pal$reference, lwd = 1.5, lty = 2)

# DPM-PM
plot(theta_true, est_comparison$dpm$theta$theta_pm,
     pch = 16, cex = 0.6, col = adjustcolor(pal$dpm_pm, 0.6),
     xlab = expression(theta[true]), ylab = expression(hat(theta)[PM]),
     main = "DPM-PM", asp = 1)
abline(0, 1, col = pal$reference, lwd = 1.5, lty = 2)

par(mfrow = c(1, 1))
```

### Loss Function Results

Three complementary loss functions capture different aspects of
estimator quality:

| Loss | Formula | Measures |
|:-----|:--------|:---------|
| **MSEL** | $\frac{1}{N}\sum_p(\hat{\theta}_p - \theta_p)^2$ | Individual accuracy (Goal 1) |
| **MSELR** | $\frac{1}{N}\sum_p(R(\hat{\theta}_p)/N - R(\theta_p)/N)^2$ | Ranking accuracy (Goal 2) |
| **KS** | $\max_t\lvert F_{\hat{\theta}}(t) - F_\theta(t)\rvert$ | Distributional fidelity (Goal 3) |

```{r loss-table, eval = has_data}
# loss_results is a list with $normal and $dpm data.frames,
# each with columns: parameter, method, msel, mselr, ks

# Combine into one table with a prior column
loss_normal <- loss_results$normal
loss_normal$prior <- "Normal"
loss_dpm <- loss_results$dpm
loss_dpm$prior <- "DPM"
loss_combined <- rbind(loss_normal, loss_dpm)

# Filter to theta and format
loss_display <- loss_combined[loss_combined$parameter == "theta", ]
loss_display <- loss_display[order(loss_display$prior, loss_display$method), ]

loss_display$Label <- paste0(loss_display$prior, "-",
                              toupper(loss_display$method))
loss_display <- loss_display[, c("Label", "msel", "mselr", "ks")]
names(loss_display) <- c("Method", "MSEL", "MSELR", "KS")
loss_display$MSEL  <- round(loss_display$MSEL, 4)
loss_display$MSELR <- round(loss_display$MSELR, 5)
loss_display$KS    <- round(loss_display$KS, 4)

knitr::kable(loss_display, row.names = FALSE,
             caption = "Loss function results: Bimodal, N=200, w-bar=0.8.",
             align = "lccc")
```

### Interpreting the Loss Table

```{r best-methods, eval = has_data}
# Identify best method for each loss
best_msel  <- loss_display$Method[which.min(loss_display$MSEL)]
best_mselr <- loss_display$Method[which.min(loss_display$MSELR)]
best_ks    <- loss_display$Method[which.min(loss_display$KS)]

cat("Best for MSEL  (individual accuracy):", best_msel, "\n")
cat("Best for MSELR (ranking accuracy):   ", best_mselr, "\n")
cat("Best for KS    (distributional):     ", best_ks, "\n")
```

### Visual Loss Comparison

```{r loss-barplot, eval = has_data, fig.height = 5, fig.width = 9, fig.cap = "Bar chart of loss values across six methods.  Lower is better for all three metrics; the best prior-method combination differs by metric."}
method_colors_vec <- c(
  "Normal-CB" = pal$normal_cb, "Normal-GR" = pal$normal_gr,
  "Normal-PM" = pal$normal_pm,
  "DPM-CB"    = pal$dpm_cb, "DPM-GR" = pal$dpm_gr,
  "DPM-PM"    = pal$dpm_pm
)

loss_long <- data.frame(
  Method = rep(loss_display$Method, 3),
  Metric = rep(c("MSEL", "MSELR", "KS"), each = nrow(loss_display)),
  Loss   = c(loss_display$MSEL, loss_display$MSELR, loss_display$KS)
)
loss_long$Method <- factor(loss_long$Method, levels = loss_display$Method)
loss_long$Metric <- factor(loss_long$Metric, levels = c("MSEL", "MSELR", "KS"))

ggplot(loss_long, aes(x = Method, y = Loss, fill = Method)) +
  geom_col(width = 0.7) +
  facet_wrap(~ Metric, scales = "free_y", ncol = 3) +
  scale_fill_manual(values = method_colors_vec, guide = "none") +
  labs(title = "Loss comparison: 6 method-prior combinations",
       y = "Loss (lower is better)") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
        strip.text = element_text(face = "bold", size = 11))
```

### Low-Reliability Contrast

The deep dive above used a 25-item test ($\bar{w} \approx 0.8$) where
the data are informative enough to compare prior and estimator choices. How
do the patterns change when reliability is much lower?

```{r lowrel-comparison, eval = (has_data && has_lowrel_data), fig.width = 10, fig.height = 5, fig.cap = "Posterior density comparison at two reliability levels (DPM prior only). Left ($\\rho \\approx 0.5$, 10 items): severe shrinkage compresses PM toward the center, while CB and GR recover more spread. Right ($\\rho \\approx 0.8$, 25 items): shrinkage is milder and the estimators move closer together."}
comp_hi <- est_comparison
comp_lo <- readRDS(find_extdata("vignette_estimates_comparison_lowrel.rds"))

build_panel <- function(comp_obj, label) {
  N <- length(comp_obj$true_theta)
  data.frame(
    value = c(comp_obj$true_theta,
              comp_obj$dpm$theta$theta_pm,
              comp_obj$dpm$theta$theta_cb,
              comp_obj$dpm$theta$theta_gr),
    Method = factor(rep(c("True", "DPM-PM", "DPM-CB", "DPM-GR"), each = N),
                    levels = c("True", "DPM-PM", "DPM-CB", "DPM-GR")),
    Reliability = label
  )
}

df_rel <- rbind(
  build_panel(comp_lo, "\u03C1 \u2248 0.5  (10 items)"),
  build_panel(comp_hi, "\u03C1 \u2248 0.8  (25 items)")
)
df_rel$Reliability <- factor(df_rel$Reliability,
  levels = c("\u03C1 \u2248 0.5  (10 items)",
             "\u03C1 \u2248 0.8  (25 items)"))

rel_colors <- c("True" = "gray40", "DPM-PM" = pal$dpm_pm,
                "DPM-CB" = pal$dpm_cb, "DPM-GR" = pal$dpm_gr)

ggplot(df_rel, aes(x = value, colour = Method, fill = Method)) +
  geom_density(alpha = 0.12, linewidth = 0.9) +
  facet_wrap(~ Reliability, ncol = 2) +
  scale_colour_manual(values = rel_colors) +
  scale_fill_manual(values = rel_colors) +
  labs(title = "Reliability determines when estimator choice matters",
       x = expression(theta), y = "Density") +
  theme_bw() +
  theme(legend.position = "top",
        strip.text = element_text(face = "bold", size = 11))
```

```{r lowrel-loss-table, eval = (has_data && has_lowrel_data), echo = FALSE}
loss_lo <- readRDS(find_extdata("vignette_loss_results_lowrel.rds"))

format_loss <- function(loss_obj, rho_label) {
  ld <- loss_obj$dpm[loss_obj$dpm$parameter == "theta",
                     c("method", "msel", "mselr", "ks")]
  ld$Reliability <- rho_label
  ld$method <- toupper(ld$method)
  ld
}

loss_both <- rbind(
  format_loss(loss_lo, "\u03C1 \u2248 0.5"),
  format_loss(loss_results, "\u03C1 \u2248 0.8")
)

knitr::kable(
  loss_both[, c("Reliability", "method", "msel", "mselr", "ks")],
  col.names = c("Reliability", "Method", "MSEL", "MSELR", "KS"),
  digits = c(0, 0, 4, 6, 4),
  row.names = FALSE,
  caption = paste("Loss values at two reliability levels.",
                  "At low reliability the gap between PM and GR on KS",
                  "widens substantially.")
)
```

> **Key observation.** At $\bar{w} \approx 0.5$, the choice of estimator
> (PM vs. CB vs. GR) matters more than the choice of prior, because every
> person's posterior is heavily shrunk toward the prior center.  CB and GR
> resist this compression, producing estimates whose distribution more
> closely tracks the truth. In this bundled low-reliability fixture, CB and
> GR also improve MSEL relative to PM, so the direction of the MSEL tradeoff
> should be checked empirically rather than assumed.

### Dispersion Recovery at Two Reliability Levels

```{r dispersion-table, eval = (has_data && has_lowrel_data), echo = FALSE}
# Compare standard deviations of estimates vs. truth
disp_stats <- function(comp_obj, rho_label) {
  true_sd <- sd(comp_obj$true_theta)
  data.frame(
    Reliability = rho_label,
    `SD(True)` = round(true_sd, 3),
    `SD(PM)` = round(sd(comp_obj$dpm$theta$theta_pm), 3),
    `SD(CB)` = round(sd(comp_obj$dpm$theta$theta_cb), 3),
    `SD(GR)` = round(sd(comp_obj$dpm$theta$theta_gr), 3),
    `PM/True` = round(sd(comp_obj$dpm$theta$theta_pm) / true_sd, 2),
    `GR/True` = round(sd(comp_obj$dpm$theta$theta_gr) / true_sd, 2),
    check.names = FALSE
  )
}

disp_both <- rbind(
  disp_stats(comp_lo, "\u03C1 \u2248 0.5"),
  disp_stats(comp_hi, "\u03C1 \u2248 0.8")
)

knitr::kable(disp_both, row.names = FALSE,
             caption = paste("Standard deviation of estimates vs. truth.",
                            "The PM/True ratio quantifies shrinkage severity;",
                            "GR/True shows how well the triple-goal estimator",
                            "recovers the true dispersion."))
```

> **Pattern.** At low reliability, PM compresses the ability distribution
> to roughly half its true spread (PM/True $\approx 0.50$), effectively
> destroying the bimodal structure.  The GR estimator retains about 70%
> of the true dispersion (GR/True $\approx 0.70$), substantially
> reducing the information loss. At high reliability, the estimators move
> closer to the true spread, though shrinkage and prior effects should still
> be checked with the relevant loss metric.

## Key Patterns

The simulation study from Lee & Wind reveals four main findings:

### Finding 1: Reliability Dominates Over Sample Size

Across all conditions, the marginal reliability $\bar{w}$ is a stronger
predictor of estimator performance than sample size $N$.
At $\bar{w} = 0.50$, prior-family differences are harder to distinguish
because the data are weak; estimator choice can still matter for
distributional loss.
At $\bar{w} = 0.90$, prior-family and estimator differences can become
easier to see.

> **Implication:** Invest in longer tests (more items) rather than
> larger samples if your goal is to discriminate between Normal and DPM
> priors.

### Finding 2: DPM Can Help Under Non-Normality + Sufficient Reliability

The DPM prior is most likely to help when:

- The true latent distribution is non-normal (bimodal or skewed), **and**
- Reliability is at least moderate ($\bar{w} \ge 0.70$).

Under normality, the DPM prior performs comparably to the Normal prior ---
the DP adapts to the true distribution and does not distort it.

### Finding 3: Match the Estimator to the Loss

The three posterior summary methods optimize different goals:

| Method | Strength | When to use |
|:-------|:---------|:------------|
| **PM** | Targets individual MSE (MSEL) | Point predictions for each person |
| **CB** | Better distributional match than PM | Group-level summary statistics |
| **GR** | Lowest KS distance | Percentile reports, distributional inference |

PM is the traditional choice for Goal 1 (minimizing per-person squared
error), though finite-sample fixtures can occasionally favor a de-shrunk
estimator on MSEL.
However, PM tends to over-shrink toward the prior center, producing
estimates whose empirical distribution is too narrow when distributional
recovery is the goal.

CB and GR "de-shrink" the estimates to better match the true distribution,
often at a cost to individual MSE, but the realized tradeoff should be read
from the loss table.

### Finding 4: No One-Size-Fits-All

No single method--prior combination dominates across all conditions and
loss functions simultaneously.
The optimal choice depends on:

- The shape of the true latent distribution (or your best guess).
- The reliability of the instrument.
- The inferential goal (individual prediction vs. distributional reporting).

> **Recommendation:** For routine use, fit both Normal and DPM models,
> compare predictive fit with WAIC, inspect posterior-summary densities, and
> select the estimator (PM, CB, or GR) that matches your reporting goals.
> WAIC evaluates item-response prediction, not ability-distribution recovery.

## Scaling to a Full Study

The single-condition pipeline above can be wrapped in a loop for a full
Monte Carlo study.
Below is a template that you would run in a non-interactive session
(e.g., on a computing cluster).
This code is shown but **not executed** in the vignette:

```{r full-study-template, eval = FALSE}
# ============================================================
# Full simulation study template
# ============================================================
# WARNING: This runs 18 conditions x 100 replications x 2 models
# Estimated time: ~72 hours on a single core
# ============================================================

library(DPMirt)

if (!requireNamespace("IRTsimrel", quietly = TRUE)) {
  stop(
    "This template uses IRTsimrel reliability targeting and the 'skew_pos' ",
    "latent-shape name. Install IRTsimrel, or replace 'skew_pos' with ",
    "'skewed' and remove reliability-targeting assumptions."
  )
}

# --- Design ---
conditions <- expand.grid(
  latent_shape = c("normal", "bimodal", "skew_pos"),
  n_persons    = c(50, 200),
  target_rho   = c(0.5, 0.7, 0.9),
  stringsAsFactors = FALSE
)

n_reps  <- 100
n_items <- 25

# --- Storage ---
all_results <- vector("list", nrow(conditions) * n_reps)
result_idx  <- 0

for (cond in seq_len(nrow(conditions))) {
  cfg <- conditions[cond, ]
  cat("Condition", cond, "/", nrow(conditions), ":",
      cfg$latent_shape, "N =", cfg$n_persons,
      "rho =", cfg$target_rho, "\n")

  for (rep in seq_len(n_reps)) {
    result_idx <- result_idx + 1
    rep_seed   <- cond * 1000 + rep

    # --- Step 1: Simulate ---
    sim <- dpmirt_simulate(
      n_persons    = cfg$n_persons,
      n_items      = n_items,
      model        = "rasch",
      target_rho   = cfg$target_rho,
      latent_shape = cfg$latent_shape,
      seed         = rep_seed
    )

    if (!identical(sim$method, "irtsimrel")) {
      stop("Expected IRTsimrel simulation for this reliability-targeted study.")
    }

    # --- Step 2: Fit Normal ---
    fit_n <- dpmirt(
      sim$response,
      model   = "rasch",
      prior   = "normal",
      niter   = 10000,
      nburnin = 2000,
      seed    = rep_seed + 1,
      verbose = FALSE
    )

    # --- Step 3: Fit DPM ---
    fit_d <- dpmirt(
      sim$response,
      model   = "rasch",
      prior   = "dpm",
      niter   = 10000,
      nburnin = 2000,
      seed    = rep_seed + 2,
      verbose = FALSE
    )

    # --- Step 4: Estimate ---
    est_n <- dpmirt_estimates(fit_n, methods = c("pm", "cb", "gr"))
    est_d <- dpmirt_estimates(fit_d, methods = c("pm", "cb", "gr"))

    # --- Step 5: Evaluate ---
    loss_n <- dpmirt_loss(est_n, true_theta = sim$theta,
                           true_beta = sim$beta)
    loss_d <- dpmirt_loss(est_d, true_theta = sim$theta,
                           true_beta = sim$beta)

    loss_n$prior <- "Normal"
    loss_d$prior <- "DPM"

    # --- Step 6: Store ---
    all_results[[result_idx]] <- cbind(
      rbind(loss_n, loss_d),
      condition    = cond,
      replication  = rep,
      latent_shape = cfg$latent_shape,
      n_persons    = cfg$n_persons,
      target_rho   = cfg$target_rho,
      achieved_rho = sim$achieved_rho,
      waic_normal  = fit_n$waic,
      waic_dpm     = fit_d$waic,
      reliability  = sim$reliability
    )
  }
}

# --- Combine ---
results_df <- do.call(rbind, all_results)

# --- Save ---
saveRDS(results_df, "simulation_results.rds")
```

### Parallelization

For faster execution, parallelize across conditions using
`parallel::mclapply()` or a cluster scheduler:

```{r parallel-template, eval = FALSE}
library(parallel)

run_one_condition <- function(cond, n_reps = 100) {
  cfg <- conditions[cond, ]
  # ... same inner loop as above ...
}

# Run on 4 cores (adjust to your machine)
results_list <- mclapply(
  seq_len(nrow(conditions)),
  run_one_condition,
  mc.cores = 4
)
```

This template assumes IRTsimrel is installed for Rasch reliability targeting
and for the `skew_pos` latent-shape name. Without IRTsimrel, DPMirt uses the
fallback simulator, ignores `target_rho`, and supports the fallback shape
names `"normal"`, `"bimodal"`, and `"skewed"`.

### Aggregation and Reporting

Once all replications are complete, aggregate results by condition:

```{r aggregate-template, eval = FALSE}
library(dplyr)

summary_table <- results_df %>%
  filter(parameter == "theta") %>%
  group_by(latent_shape, n_persons, target_rho, prior, method) %>%
  summarise(
    mean_msel  = mean(msel),
    se_msel    = sd(msel) / sqrt(n()),
    mean_mselr = mean(mselr),
    mean_ks    = mean(ks),
    .groups    = "drop"
  )
```

## Extending the Study

The DPMirt simulation framework supports several extensions beyond the
basic Rasch study:

### 2PL Models

```{r extend-2pl, eval = FALSE}
sim_2pl <- dpmirt_simulate(
  n_persons    = 200,
  n_items      = 25,
  model        = "2pl",
  target_rho   = 0.8,
  latent_shape = "bimodal",
  seed         = 42
)

fit_2pl <- dpmirt(
  sim_2pl$response,
  model = "2pl",
  prior = "dpm",
  niter = 15000,
  nburnin = 3000
)
```

### Custom Latent Distributions

With IRTsimrel installed, you have access to 12 distribution shapes:

```{r latent-shapes, eval = FALSE}
# Built-in supported shapes (when IRTsimrel is available)
shapes <- c("normal", "bimodal", "trimodal", "multimodal",
            "skew_pos", "skew_neg", "heavy_tail", "light_tail",
            "uniform", "floor", "ceiling")

for (shape in shapes) {
  sim <- dpmirt_simulate(200, 25, latent_shape = shape, seed = 42)
  cat(shape, ": KR-20 =", round(sim$reliability, 3), "\n")
}

# Custom shapes require IRTsimrel, a Rasch or 2PL simulation,
# and latent_params = list(mixture_spec = ...).
custom_sim <- dpmirt_simulate(
  200, 25,
  latent_shape  = "custom",
  latent_params = list(mixture_spec = list(
    weights = c(0.5, 0.5),
    means   = c(-1, 1),
    sds     = c(0.5, 0.5)
  )),
  seed          = 42
)
```

Custom latent distributions are not available in fallback simulation or for
3PL.

### Custom Loss Functions

Extend evaluation with your own loss function:

```{r custom-loss, eval = FALSE}
# Bias (signed error)
bias_loss <- function(estimate, true) mean(estimate - true)

# Coverage of 95% CI
coverage_loss <- function(estimate, true) {
  # Requires access to full posterior...
  # Use dpmirt_draws() for this
}

loss_custom <- dpmirt_loss(
  est_dpm,
  true_theta  = sim$theta,
  custom_loss = bias_loss
)
```

## Diagnostic Checks for Simulation Validity

Before interpreting results, verify that the simulation and MCMC
converged properly.

### Check 1: Achieved Reliability

```{r check-reliability}
cat("Target reliability:", 0.8, "\n")
cat("Empirical KR-20: ", round(sim$reliability, 4), "\n")
if (!is.null(sim$achieved_rho)) {
  cat("Calibration rho: ", round(sim$achieved_rho, 4), "\n")
}
cat("Simulation method: ", sim$method, "\n")
```

### Check 2: MCMC Convergence

For pre-computed fits, you would check:

```r
# Minimum ESS across all parameters
min(fit$ess$items)   # Review relative to retained thinned draws
min(fit$ess$theta)

# Trace plot
plot(fit, type = "trace")

# WAIC comparison
cat("WAIC Normal:", fit_normal$waic, "\n")
cat("WAIC DPM:   ", fit_dpm$waic, "\n")
```

For the bundled compact fixtures, ESS is computed from 250 retained thinned
draws and is intended for documentation-scale screening. Use full simulation
runs, with many more retained draws, for convergence assessment in a study.

### Check 3: Posterior Predictive Check

```r
# Compare observed vs. predicted sum score distribution
plot(fit, type = "pp_check")
```

## Summary Table of Functions

| Function | Purpose | Phase |
|:---------|:--------|:------|
| `dpmirt_simulate()` | Generate IRT data; target reliability for Rasch/2PL when IRTsimrel is available | Simulation |
| `dpmirt()` | Fit model (one-step) | Fitting |
| `dpmirt_estimates()` | Compute PM, CB, GR estimates | Estimation |
| `dpmirt_loss()` | Evaluate MSEL, MSELR, KS | Evaluation |
| `dpmirt_draws()` | Extract retained rescaled posterior draws | Post-hoc analysis |
| `dpmirt_dp_density()` | Compute DP mixture density | DPM diagnostics |

## What's Next?

| Vignette | Why read it |
|:---------|:------------|
| [Quick Start](quick-start.html) | Simpler walkthrough of the basic pipeline |
| [Prior Elicitation](prior-elicitation.html) | Optional DPprior calibration of $\alpha$ and Gamma(1,3) fallback |
| [NIMBLE Internals](nimble-internals.html) | NIMBLE internals, live compile-once pattern, custom sampler hooks |

## References

Lee, J. & Wind, S. Targeting toward inferential goals in Bayesian Rasch
models for estimating person-specific latent traits. *OSF Preprint*.
<https://doi.org/10.31219/osf.io/qrw4n>

Lee, J. (2025). Reliability-targeted simulation of item response data:
Solving the inverse design problem. arXiv preprint arXiv:2512.16012.
https://arxiv.org/abs/2512.16012

Paganin, S., Paciorek, C. J., Wehrhahn, C., Rodríguez, A.,
Rabe-Hesketh, S., & de Valpine, P. (2023). Computational strategies
and estimation performance with Bayesian semiparametric item response
theory models. *Journal of Educational and Behavioral Statistics*,
48(2), 147--188. https://doi.org/10.3102/10769986221136105

Ghosh, M. (1992). Constrained Bayes estimation with applications.
*JASA*, 87(418), 533--540.

Shen, W., & Louis, T. A. (1998). Triple-goal estimates in two-stage
hierarchical models. *JRSS-B*, 60(2), 455--471.
